---
title: "Food Prices in Pakistan"
subtitle: "A complete analytical workflow with dplyr 1.2.0, ggplot2 and sf"
author: "Prof. Dr. Zahid Asghar"
date: today
format:
  html:
    theme: sketchy
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-copy: true
    code-overflow: wrap
    df-print: paged
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 200
    embed-resources: true
execute:
  # Set to `true` when you are ready to render live against the HDX API.
  # Kept `false` so the document renders on a machine with no internet.
  eval: false
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
---

::: {.callout-note appearance="simple"}
## How to use this tutorial

Work through it **in order** and **type the code** rather than pasting it. Every
section ends with a *Check* — a small piece of code whose only job is to tell you
whether the previous step did what you thought it did. In real analytical work
those checks are the difference between a defensible number and an embarrassing one.

The dataset is the WFP VAM Food Prices series for Pakistan, published on the
Humanitarian Data Exchange. It is monthly, market-level, and runs for roughly two
decades — which makes it one of the few genuinely long, genuinely public,
genuinely sub-national price series available for Pakistan.
:::

# What you will be able to do by the end

1. Pull a live dataset from an API reproducibly, and cache it so you are not at the
   mercy of the network in the middle of a workshop.
2. Recognise and strip a HXL header row, and understand why the file must first be
   read as all-character.
3. Use the **dplyr 1.2.0** filtering and recoding families (`filter_out()`,
   `when_any()`, `when_all()`, `recode_values()`, `replace_values()`,
   `replace_when()`) and — more importantly — know *which one* to reach for.
4. Make the four analytical decisions that determine whether your "average food
   price" number is meaningful at all.
5. Build inflation, real-price, affordability, seasonality and market-integration
   measures from raw price observations.
6. Draw a publication-quality provincial choropleth of Pakistan that includes
   Azad Jammu & Kashmir, Gilgit-Baltistan and the wider Jammu & Kashmir region,
   with the disputed area shown as *no data* and a proper cartographic disclaimer.

# Setup

## Packages

```{r}
#| label: setup
library(tidyverse)        # dplyr >= 1.2.0, ggplot2, stringr, purrr, tidyr, readr
library(sf)               # simple features: spatial data as data frames
library(rnaturalearth)    # boundary polygons
library(scales)           # axis label formatting
library(here)             # project-relative paths
library(jsonlite)         # for the HDX API call

theme_set(theme_minimal(base_size = 12))
options(pillar.sigfig = 4)
```

`rnaturalearth` needs a companion package for high-resolution (scale 10) admin-1
boundaries. It is not on CRAN:

```{r}
#| label: install-hires
# Run once
install.packages("rnaturalearthhires", repos = "https://ropensci.r-universe.dev")
library(rnaturalearthhires)
```

## Version check — fail loudly, early

Half of this tutorial uses functions that did not exist before dplyr 1.2.0
(released February 2026). Check now rather than discovering it in section 4.

```{r}
#| label: version-check
if (packageVersion("dplyr") < "1.2.0") {
  stop(
    "This tutorial needs dplyr >= 1.2.0; you have ", packageVersion("dplyr"),
    ". Run: install.packages('dplyr')"
  )
}
```

::: {.callout-tip}
## If you are stuck on an older dplyr

Public-sector laptops are often frozen at an older R. Every new verb has a
pre-1.2.0 equivalent — uglier, but functionally identical. Keep this table handy:

| dplyr 1.2.0 | Pre-1.2.0 equivalent |
|---|---|
| `filter_out(x == "a")` | `filter(!(x == "a") \| is.na(x))` |
| `filter(when_any(a, b))` | `filter(a \| b)` (plus NA handling) |
| `filter(when_all(a, b))` | `filter(a, b)` |
| `recode_values(x, "a" ~ 1, "b" ~ 2)` | `case_match(x, "a" ~ 1, "b" ~ 2)` |
| `replace_values(x, "a" ~ "A")` | `case_match(x, "a" ~ "A", .default = x)` |
| `replace_when(x, cond ~ v)` | `if_else(cond, v, x)` |

Note that `case_match()` is **deprecated** as of dplyr 1.2.0 in favour of
`recode_values()` — so the migration runs in one direction only.
:::

# Getting the data reproducibly

## Why you cannot just `read_csv()` the web page

`https://data.humdata.org/dataset/wfp-food-prices-for-pakistan` is a *landing
page*. Pointing `read_csv()` at it downloads HTML. HDX runs on CKAN, which exposes
a JSON API, and the API is the stable, scriptable way in.

```{r}
#| label: hdx-lookup
hdx_resource_url <- function(dataset_id,
                             include = "food_prices",
                             exclude = "qc") {
  pkg <- fromJSON(
    paste0("https://data.humdata.org/api/3/action/package_show?id=", dataset_id)
  )

  resources <- as_tibble(pkg$result$resources)

  resources |>
    filter(str_detect(url, regex(include, ignore_case = TRUE))) |>
    filter_out(str_detect(url, regex(exclude, ignore_case = TRUE))) |>
    slice(1) |>
    pull(url)
}
```

Two things to notice in that pipeline:

- The **positive** condition uses `filter()`, the **negative** one uses
  `filter_out()`. That is the whole design intent of the new verb: the function
  name, not a buried `!`, tells the reader whether rows are being kept or dropped.
- HDX publishes a "QuickCharts" (`qc`) subset alongside the full file. It has the
  same name pattern and will silently give you a truncated dataset if you take the
  first match blindly. This is a very common and very quiet failure mode.

## Cache the download

Never re-download in a loop, in a render, or in front of thirty participants on
conference wifi.

```{r}
#| label: download-cache
raw_dir  <- here("data", "raw")
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)

wfp_path <- file.path(raw_dir, "wfp_food_prices_pak.csv")

if (!file.exists(wfp_path)) {
  wfp_url <- hdx_resource_url("wfp-food-prices-for-pakistan")
  message("Downloading from: ", wfp_url)
  download.file(wfp_url, destfile = wfp_path, mode = "wb")
} else {
  message("Using cached file: ", wfp_path)
}
```

::: {.callout-important}
## Record the vintage

A live API means your results change when WFP revises the file. For anything that
will be published, record *when* you pulled it and keep the raw file under version
control (or at least in a dated folder). Reproducibility is not the same as
re-runnability.

```{r}
#| label: vintage
vintage <- file.info(wfp_path)$mtime
vintage
```
:::

## Read everything as character — deliberately

```{r}
#| label: read-raw
raw <- read_csv(wfp_path, col_types = cols(.default = col_character()))

dim(raw)
glimpse(raw)
```

Why force character? Because row 1 is not data:

```{r}
#| label: hxl-row
raw |> slice(1) |> select(1:6) |> glimpse()
```

Every HDX file carries a **HXL tag row** — `#date`, `#adm1+name`, `#value` and so
on — as the first data row. It is a machine-readable schema annotation, not an
observation. If you let `read_csv()` guess column types, it sees `#date` in a date
column, decides the whole column is character anyway, and you gain nothing but
lose control. Reading as character and converting *after* removing the HXL row is
explicit and predictable.

# Cleaning

## Drop the HXL row

```{r}
#| label: drop-hxl
prices_raw <- raw |>
  filter_out(str_starts(date, "#"))

nrow(raw) - nrow(prices_raw)   # should be exactly 1
```

## The single most important behaviour in this tutorial

`filter()` and `filter_out()` **both treat `NA` like `FALSE`**. Read that twice.
For `filter()`, `NA` means *not kept*. For `filter_out()`, `NA` means *not
dropped*. That asymmetry is precisely what makes `filter_out()` safe.

```{r}
#| label: na-demo
demo <- tibble(
  market   = c("Lahore", "Karachi", "Quetta", "Peshawar"),
  category = c("cereals and tubers", "non-food", NA, "pulses and nuts")
)

# 1. The naive negation — silently loses Quetta
demo |> filter(category != "non-food")

# 2. The defensive negation — correct, but you have to remember to write it
demo |> filter(category != "non-food" | is.na(category))

# 3. filter_out() — correct by construction
demo |> filter_out(category == "non-food")
```

::: {.callout-warning}
## This is not a toy problem

In the WFP file, `category` is missing for a non-trivial number of rows. Version 1
of that pipeline throws away real observations and gives you a slightly-too-low
national average, with no warning and no error. That kind of bug survives peer
review because the output *looks* fine.
:::

## Type conversion

```{r}
#| label: clean-types
prices <- prices_raw |>
  mutate(
    date = ymd(date),
    across(c(latitude, longitude, price, usdprice), as.numeric),
    across(where(is.character), str_squish),
    year      = year(date),
    month_num = month(date),
    month_lab = month(date, label = TRUE, abbr = TRUE)
  )
```

**Check.** Silent coercion failures produce `NA`, so count them:

```{r}
#| label: check-coercion
prices |>
  summarise(
    bad_date  = sum(is.na(date)),
    bad_price = sum(is.na(price)),
    bad_lat   = sum(is.na(latitude))
  )
```

If `bad_date` is anything other than 0, stop and look at the offending values
before going further — do not "fix" it downstream.

## Missingness audit

```{r}
#| label: missingness
prices |>
  summarise(across(everything(), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  mutate(pct_missing = n_missing / nrow(prices)) |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

## Know your columns

The WFP VAM schema is stable across countries. Learn it once and every other
country file is free.

| Column | Meaning | Why it matters |
|---|---|---|
| `date` | Reference month (day is always 15) | Monthly, not daily |
| `admin1`, `admin2` | Province, district | Your join keys to geography |
| `market` | Reporting market | The actual unit of observation |
| `latitude`, `longitude` | Market coordinates | Point mapping, distance calcs |
| `category` | Broad food group; also `non-food` | Fuel and wages live here |
| `commodity` | Item, e.g. *Wheat flour* | Not standardised across countries |
| `unit` | KG, L, 20 KG, Unit, Day… | **Must be normalised** |
| `pricetype` | Retail / Wholesale | **Never mix these** |
| `priceflag` | actual / aggregate | `aggregate` are WFP roll-ups |
| `currency` | PKR | Constant here |
| `price` | Price in local currency | Nominal |
| `usdprice` | Same price at that month's FX rate | Nominal USD, *not* real |

```{r}
#| label: schema-counts
prices |> count(pricetype, priceflag, sort = TRUE)
prices |> count(category, sort = TRUE)
prices |> count(unit, sort = TRUE)
prices |> count(admin1, sort = TRUE)
prices |> count(market, sort = TRUE) |> print(n = Inf)
```

# The four decisions that make or break the analysis

Everything above was mechanical. Everything below is a judgement call, and each one
changes your headline number. Make them explicitly and write them down.

## Decision 1 — `actual` versus `aggregate`

`aggregate` rows are WFP-computed roll-ups over other rows. Keeping them
double-counts.

```{r}
#| label: d1-priceflag
prices_actual <- prices |>
  filter_out(priceflag != "actual")

nrow(prices) - nrow(prices_actual)
```

Note the shape of that condition. `filter_out(priceflag != "actual")` drops
everything that is *not* actual, and — because `NA` is treated as `FALSE` — keeps
rows where `priceflag` is missing. Whether that is what you want is a substantive
choice; the point is that the code makes it visible.

## Decision 2 — retail versus wholesale

This is the decision most often missed. Retail and wholesale prices for the same
commodity in the same month differ by a margin. If you average across both, your
"national price" moves whenever the *mix* of reporting types changes — even if no
actual price moved.

```{r}
#| label: d2-pricetype
prices_actual |>
  count(pricetype, commodity) |>
  pivot_wider(names_from = pricetype, values_from = n, values_fill = 0) |>
  arrange(desc(Retail))
```

```{r}
#| label: d2-retail
retail <- prices_actual |>
  filter(pricetype == "Retail")
```

Analyse wholesale separately if you need it. Do not pool.

## Decision 3 — normalise the units

`price` is per `unit`, and `unit` is a free-text mess: `KG`, `20 KG`, `500 G`,
`L`, `Unit`, `Day`. A mean over mixed units is meaningless. Write a small,
testable function.

```{r}
#| label: d3-unit-fns
#' Convert a WFP unit string to its kilogram equivalent
#' Returns NA for units that are not mass-based.
unit_to_kg <- function(unit) {
  qty  <- coalesce(parse_number(unit), 1)      # "20 KG" -> 20; "KG" -> 1
  mult <- case_when(
    str_detect(unit, regex("\\bMT\\b",  ignore_case = TRUE)) ~ 1000,
    str_detect(unit, regex("\\bKG\\b",  ignore_case = TRUE)) ~ 1,
    str_detect(unit, regex("\\bG\\b",   ignore_case = TRUE)) ~ 1 / 1000,
    .default = NA_real_
  )
  qty * mult
}

#' Convert a WFP unit string to its litre equivalent
unit_to_litre <- function(unit) {
  qty  <- coalesce(parse_number(unit), 1)
  mult <- case_when(
    str_detect(unit, regex("\\bML\\b", ignore_case = TRUE)) ~ 1 / 1000,
    str_detect(unit, regex("\\bL\\b",  ignore_case = TRUE)) ~ 1,
    .default = NA_real_
  )
  qty * mult
}
```

Order matters in `case_when()` — the first match wins. `MT` is tested before `KG`
and `ML` before `L` for exactly that reason. The `\\b` word boundaries stop `G`
from matching inside `KG`.

**Test the function before trusting it.** This is a habit worth drilling:

```{r}
#| label: d3-test
test_units <- c("KG", "20 KG", "500 G", "MT", "L", "1.5 L", "Unit", "Day", "Dozen")

tibble(
  unit  = test_units,
  kg    = unit_to_kg(test_units),
  litre = unit_to_litre(test_units)
)
```

Now apply and audit what did not convert:

```{r}
#| label: d3-apply
retail <- retail |>
  mutate(
    kg_equiv     = unit_to_kg(unit),
    l_equiv      = unit_to_litre(unit),
    price_per_kg = price / kg_equiv,
    price_per_l  = price / l_equiv,
    price_std    = coalesce(price_per_kg, price_per_l),
    unit_std     = case_when(
      !is.na(price_per_kg) ~ "per kg",
      !is.na(price_per_l)  ~ "per litre",
      .default = NA_character_
    )
  )

# Everything the function could not classify — inspect, don't ignore
retail |>
  filter(is.na(unit_std)) |>
  count(category, commodity, unit, sort = TRUE)
```

You should see wages (`Day`), fuel (`L` — caught, but non-food), eggs (`Dozen`)
and similar. Nothing food-and-mass-based should appear. If it does, extend the
function.

## Decision 4 — the panel is unbalanced

This is the subtlest one. Markets enter and leave the sample. A simple
`mean(price)` across whatever markets reported this month is a **composition-varying**
statistic: if an expensive market starts reporting in March, your national average
jumps in March with no price change anywhere.

```{r}
#| label: d4-coverage
coverage <- retail |>
  filter(!is.na(price_std)) |>
  summarise(
    n_obs     = n(),
    n_months  = n_distinct(date),
    first_obs = min(date),
    last_obs  = max(date),
    .by = c(commodity, market)
  ) |>
  arrange(commodity, market)

coverage
```

```{r}
#| label: d4-coverage-plot
retail |>
  filter(str_detect(commodity, regex("wheat flour", ignore_case = TRUE))) |>
  distinct(market, date) |>
  ggplot(aes(date, fct_rev(fct_infreq(market)))) +
  geom_point(size = 0.6, alpha = 0.7) +
  labs(
    title    = "Who reports, and when",
    subtitle = "Wheat flour, retail — each dot is one market-month",
    x = NULL, y = NULL,
    caption  = "Gaps and late entries are why a raw cross-market mean is unsafe"
  )
```

Look at that plot before you compute a single average. We return to the fix in
@sec-index.

# The dplyr 1.2.0 verbs, properly {#sec-verbs}

## `when_any()` and `when_all()`

`filter()` combines its conditions with `&`. To get `|` you used to need
parentheses and one long unreadable expression. `when_any()` and `when_all()`
let you build the boolean structure as *nested, named* pieces.

Our staple basket: food items priced per kilo, plus cooking oils and ghee priced
per litre.

```{r}
#| label: basket
basket <- retail |>
  filter_out(category == "non-food") |>
  filter(
    when_any(
      when_all(unit_std == "per kg",    category != "milk and dairy"),
      when_all(unit_std == "per litre", str_detect(commodity, regex("oil|ghee", ignore_case = TRUE)))
    )
  )

basket |> count(commodity, unit_std, sort = TRUE)
```

Read that aloud: *keep a row if (it is per-kg and not dairy) or (it is per-litre
and it is an oil or ghee)*. The code now has the same shape as the sentence.
`when_any()` and `when_all()` are ordinary vector functions — you can use them in
`mutate()`, `if_else()`, anywhere.

## Choosing among the four recoding verbs

This is the decision table. Pin it to the wall.

| You want to… | Match on | Use |
|---|---|---|
| Build a **new** column from **conditions** | conditions | `case_when()` |
| Build a **new** column from **values** (a lookup) | values | `recode_values()` |
| **Update some** values of an existing column, by condition | conditions | `replace_when()` |
| **Update some** values of an existing column, by value | values | `replace_values()` |

The critical difference is what happens to the rows you did *not* mention:

- `recode_values()` and `case_when()` **discard** them (→ `NA`, or `default`).
  They are *total* mappings.
- `replace_values()` and `replace_when()` **keep** them unchanged.
  They are *partial* mappings.

::: {.callout-warning}
## The bug this table prevents

```{r}
#| label: recode-trap
# WRONG if there is any admin1 you forgot — Islamabad, AJK, Gilgit-Baltistan
# all become NA, silently.
basket |>
  mutate(province = recode_values(
    admin1,
    "Punjab"             ~ "Punjab",
    "Sindh"              ~ "Sindh",
    "Khyber Pakhtunkhwa" ~ "KP",
    "Balochistan"        ~ "Balochistan"
  )) |>
  count(admin1, province)
```

Run that and check the `province` column for `NA`. Any dataset with more than four
admin-1 units will bite you here.
:::

## The lookup-table interface

For anything with more than a handful of cases, `recode_values()` accepts
`from`/`to` vectors instead of formulas. A `tribble()` where a cell holds a
*vector* becomes a list column, and `recode_values()` treats each vector as "any
of these map to this". This is the cleanest way to handle spelling variants.

```{r}
#| label: province-lookup
province_lookup <- tribble(
  ~from,                                                              ~to,
  c("Punjab"),                                                        "Punjab",
  c("Sindh"),                                                         "Sindh",
  c("Khyber Pakhtunkhwa", "N.W.F.P.", "NWFP",
    "Federally Administered Tribal Areas", "F.A.T.A."),               "KP",
  c("Balochistan", "Baluchistan"),                                    "Balochistan",
  c("Islamabad", "F.C.T.", "Islamabad Capital Territory"),            "Islamabad",
  c("Azad Kashmir", "Azad Jammu and Kashmir", "AJK"),                 "Azad Jammu & Kashmir",
  c("Gilgit-Baltistan", "Northern Areas"),                            "Gilgit-Baltistan"
)

province_lookup
province_lookup$from   # note: a list column
```

One lookup, used twice — once on the price data, once on the map. That is the
whole point: the crosswalk lives in exactly one place.

```{r}
#| label: apply-lookup
basket <- basket |>
  mutate(
    province = recode_values(
      admin1,
      from = province_lookup$from,
      to   = province_lookup$to
    )
  )

# Audit: anything unmatched?
basket |>
  filter(is.na(province), !is.na(admin1)) |>
  count(admin1, sort = TRUE)
```

If that audit returns rows, add them to the lookup. Once it is empty, switch on
strict mode so the pipeline fails loudly if WFP ever renames a province:

```{r}
#| label: strict-lookup
basket |>
  filter(!is.na(admin1)) |>
  mutate(province = recode_values(
    admin1,
    from      = province_lookup$from,
    to        = province_lookup$to,
    unmatched = "error"
  )) |>
  invisible()
```

## `replace_when()` — surgical edits to one column

`replace_when()` is `base::replace()` with a grammar. Use it to winsorise
implausible spikes without touching anything else:

```{r}
#| label: winsorise
basket <- basket |>
  mutate(
    price_w = replace_when(
      price_std,
      price_std > quantile(price_std, 0.999, na.rm = TRUE)
        ~ quantile(price_std, 0.999, na.rm = TRUE)
    ),
    .by = c(commodity, unit_std)
  )

# What actually changed?
basket |>
  filter(price_std != price_w) |>
  select(date, market, commodity, unit_std, price_std, price_w) |>
  arrange(desc(price_std))
```

Always print the rows you changed. Winsorising is a defensible choice; doing it
invisibly is not.

::: {.callout-tip}
## Why `replace_when()` and not `case_when()`

`case_when()` would force you to write a second branch — `.default = price_std` —
to preserve the untouched values, and it would let you accidentally change the
column's type. `replace_when()` guarantees the output has the same type as the
input and preserves everything you did not name.
:::

# Building the analysis frames

Give every derived table a name and a clear grain (one row per *what*?). Stating
the grain in a comment is the cheapest bug-prevention in data work.

```{r}
#| label: frames
# Grain: one row per market
market_meta <- basket |>
  distinct(market, admin1, province, latitude, longitude)

# Check the grain actually holds — duplicate coordinates for one market are common
market_meta |> count(market, sort = TRUE) |> filter(n > 1)
```

If that returns rows, the same market name carries more than one coordinate pair.
Resolve it explicitly rather than letting a join silently multiply your data:

```{r}
#| label: fix-market-grain
market_meta <- basket |>
  summarise(
    latitude  = median(latitude,  na.rm = TRUE),
    longitude = median(longitude, na.rm = TRUE),
    .by = c(market, admin1, province)
  ) |>
  distinct(market, .keep_all = TRUE)
```

```{r}
#| label: frames-2
# Grain: one row per commodity-month (national)
national_monthly <- basket |>
  summarise(
    price_pkr = mean(price_w,  na.rm = TRUE),
    price_usd = mean(usdprice, na.rm = TRUE),
    n_markets = n_distinct(market),
    .by = c(date, commodity, unit_std)
  ) |>
  arrange(commodity, date)

# Grain: one row per province-commodity-year
province_year <- basket |>
  summarise(
    price_pkr = mean(price_w, na.rm = TRUE),
    n_markets = n_distinct(market),
    .by = c(province, commodity, year)
  )
```

## Pick your reference commodity from the data, not from memory

Commodity labels are country-specific and change over time. Never hard-code a
string you have not seen in the file.

```{r}
#| label: find-commodities
basket |>
  distinct(commodity, unit_std) |>
  filter(str_detect(commodity, regex("wheat|rice|sugar|oil|ghee|lentil|pulse",
                                     ignore_case = TRUE))) |>
  arrange(commodity) |>
  print(n = Inf)
```

Copy the exact string you want from that output into the next chunk.

```{r}
#| label: wheat
wheat <- basket |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)))

# Grain: one row per market-month
wheat_market <- wheat |>
  summarise(price_pkr = mean(price_w, na.rm = TRUE), .by = c(date, market, province))

# Grain: one row per month
wheat_national <- national_monthly |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)))
```

# Analysis I — Inflation

## Month-on-month and year-on-year

```{r}
#| label: inflation
national_monthly <- national_monthly |>
  arrange(commodity, date) |>
  mutate(
    mom = price_pkr / lag(price_pkr)      - 1,
    yoy = price_pkr / lag(price_pkr, 12)  - 1,
    .by = commodity
  )
```

::: {.callout-warning}
## `lag()` counts rows, not months

`lag(price_pkr, 12)` gives you the price twelve *rows* earlier. If a commodity has
a reporting gap, that is not twelve months earlier — and your "year-on-year"
number is quietly wrong. Make the panel complete first.
:::

```{r}
#| label: inflation-safe
national_monthly <- national_monthly |>
  complete(commodity, date) |>          # insert explicit NA rows for missing months
  arrange(commodity, date) |>
  mutate(
    mom = price_pkr / lag(price_pkr)     - 1,
    yoy = price_pkr / lag(price_pkr, 12) - 1,
    .by = commodity
  )
```

Now a gap produces an honest `NA` instead of a fabricated growth rate. The habit
generalises: **any time you use `lag()`, `lead()`, `diff()` or `cumsum()` on a
panel, complete the panel first.**

## Where were the big moves?

```{r}
#| label: big-moves
national_monthly |>
  filter(!is.na(yoy)) |>
  slice_max(yoy, n = 15) |>
  select(date, commodity, price_pkr, yoy)
```

# Analysis II — Nominal, USD, and real

`usdprice` is *not* a real price. It is the same nominal price converted at that
month's exchange rate. The gap between the PKR and USD series is the exchange rate
story, which is interesting but is not deflation.

```{r}
#| label: pkr-vs-usd
wheat_national |>
  select(date, price_pkr, price_usd) |>
  pivot_longer(c(price_pkr, price_usd), names_to = "series", values_to = "value") |>
  mutate(series = recode_values(series,
    "price_pkr" ~ "Nominal PKR/kg",
    "price_usd" ~ "Nominal USD/kg"
  )) |>
  ggplot(aes(date, value)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ series, scales = "free_y") +
  labs(
    title    = "Wheat flour: local currency versus US dollars",
    subtitle = "The difference between these panels is exchange rate movement, not inflation",
    x = NULL, y = NULL
  )
```

## A genuine real price

To deflate you need a price index. World Bank WDI carries annual CPI for Pakistan;
PBS publishes monthly CPI, which is better if you can get it into R.

```{r}
#| label: real-prices
cpi <- WDI::WDI(
  country   = "PK",
  indicator = c(cpi = "FP.CPI.TOTL"),
  start     = 2004,
  end       = as.integer(format(Sys.Date(), "%Y"))
) |>
  as_tibble() |>
  select(year, cpi) |>
  filter_out(is.na(cpi))

base_year <- cpi |> slice_max(year) |> pull(year)
base_cpi  <- cpi |> filter(year == base_year) |> pull(cpi)

wheat_real <- wheat_national |>
  summarise(price_pkr = mean(price_pkr, na.rm = TRUE), .by = year) |>
  inner_join(cpi, by = join_by(year)) |>
  mutate(price_real = price_pkr * base_cpi / cpi)

wheat_real |>
  pivot_longer(c(price_pkr, price_real), names_to = "series", values_to = "value") |>
  mutate(series = recode_values(series,
    "price_pkr"  ~ "Nominal",
    "price_real" ~ paste0("Real (", base_year, " prices)")
  )) |>
  ggplot(aes(year, value, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(labels = label_number(prefix = "Rs ")) +
  labs(
    title    = "Wheat flour, nominal versus real",
    subtitle = "Deflated by headline CPI",
    x = NULL, y = "Rs per kg", colour = NULL
  )
```

::: {.callout-note}
## Interpret this carefully

Headline CPI *contains* food. Deflating a food price by headline CPI gives you the
price of wheat flour **relative to the general basket** — a relative price, not a
purchasing-power-adjusted price. That is usually what you want for a food security
argument, but say so explicitly in your write-up.
:::

# Analysis III — Affordability

A price is only meaningful against an income. The same WFP file contains a daily
wage series under `non-food` — which is exactly why we did not throw those rows
away at the start.

```{r}
#| label: wages
wages <- prices_actual |>
  filter(str_detect(commodity, regex("wage", ignore_case = TRUE))) |>
  summarise(wage_pkr = mean(price, na.rm = TRUE), .by = c(date, market))

wages |> count(market, sort = TRUE)
```

```{r}
#| label: affordability
affordability <- wheat_market |>
  inner_join(wages, by = join_by(date, market)) |>
  mutate(kg_per_day_wage = wage_pkr / price_pkr)

affordability |>
  summarise(kg_per_day_wage = mean(kg_per_day_wage, na.rm = TRUE),
            n_markets = n_distinct(market),
            .by = date) |>
  ggplot(aes(date, kg_per_day_wage)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, colour = "firebrick") +
  labs(
    title    = "How much wheat flour does a day of unskilled labour buy?",
    subtitle = "Kilograms per daily wage, mean across reporting markets",
    x = NULL, y = "kg per day's wage",
    caption  = "Source: WFP VAM Food Prices Database via HDX"
  )
```

This single chart is often more persuasive to a policy audience than any price
series, because it answers the question they actually have. Note the `inner_join()`:
it silently restricts you to market-months where *both* a wheat price and a wage
were reported. Check how much that costs you:

```{r}
#| label: join-cost
nrow(wheat_market)
nrow(affordability)
```

# Analysis IV — Seasonality

A raw boxplot by calendar month conflates seasonality with the trend: if prices
rise steadily and your sample starts in March, March looks cheap. Strip the trend
first by including a year effect.

```{r}
#| label: seasonality-raw
wheat |>
  ggplot(aes(month_lab, price_w)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.4) +
  labs(title = "Raw monthly distribution — trend and season confounded",
       x = NULL, y = "Rs per kg")
```

```{r}
#| label: seasonality-model
library(marginaleffects)

season_data <- wheat_national |>
  filter(!is.na(price_pkr)) |>
  mutate(
    year_f  = factor(year(date)),
    month_f = factor(month(date, label = TRUE, abbr = TRUE), ordered = FALSE)
  )

fit_season <- lm(log(price_pkr) ~ year_f + month_f, data = season_data)

season_effects <- avg_predictions(fit_season, variables = "month_f") |>
  as_tibble() |>
  mutate(across(c(estimate, conf.low, conf.high), exp))

season_effects |>
  ggplot(aes(month_f, estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  labs(
    title    = "Wheat flour seasonality, net of year effects",
    subtitle = "Predicted price by calendar month, holding the year fixed",
    x = NULL, y = "Rs per kg"
  )
```

Reading it: the pattern should track the wheat harvest and the release of
government stocks. If it does not, that is a finding worth chasing, not a bug to
paper over.

# Analysis V — Market integration {#sec-index}

## Dispersion across markets

If markets are well integrated, arbitrage keeps prices close. The spread and the
coefficient of variation across markets are simple, defensible diagnostics.

```{r}
#| label: dispersion
dispersion <- wheat_market |>
  filter(!is.na(price_pkr)) |>
  summarise(
    n_markets = n(),
    cheapest  = min(price_pkr),
    dearest   = max(price_pkr),
    spread    = dearest - cheapest,
    cv        = sd(price_pkr) / mean(price_pkr),
    .by = date
  ) |>
  filter(n_markets >= 5)      # a spread over 2 markets is not a spread

dispersion |> slice_max(spread, n = 10)
```

The `n_markets >= 5` filter matters. Both `min()` and `max()` are order statistics:
they mechanically widen as the sample grows. Comparing a 3-market spread with a
20-market spread tells you about coverage, not integration. The CV is more robust,
which is why it is worth reporting alongside.

## Does distance explain price gaps?

The law of one price says that markets further apart should show bigger price gaps,
bounded by transport costs. This is testable with what you already have.

```{r}
#| label: distances
markets_sf <- market_meta |>
  filter_out(is.na(latitude) | is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

dist_km <- st_distance(markets_sf) |>
  units::set_units("km") |>
  units::drop_units()

dimnames(dist_km) <- list(markets_sf$market, markets_sf$market)

pair_distance <- dist_km |>
  as_tibble(rownames = "market_a") |>
  pivot_longer(-market_a, names_to = "market_b", values_to = "distance_km") |>
  filter(market_a < market_b)
```

```{r}
#| label: pair-gaps
pair_gaps <- wheat_market |>
  select(date, market, price_pkr) |>
  inner_join(
    wheat_market |> select(date, market, price_pkr),
    by            = join_by(date),
    relationship  = "many-to-many",
    suffix        = c("_a", "_b")
  ) |>
  filter(market_a < market_b) |>
  mutate(gap = abs(price_pkr_a - price_pkr_b)) |>
  summarise(
    mean_gap = mean(gap, na.rm = TRUE),
    n_months = n(),
    .by = c(market_a, market_b)
  ) |>
  filter(n_months >= 24) |>
  inner_join(pair_distance, by = join_by(market_a, market_b))

pair_gaps |>
  ggplot(aes(distance_km, mean_gap)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.6) +
  labs(
    title    = "Law of one price: do distant markets diverge?",
    subtitle = "Mean absolute wheat flour price gap against great-circle distance",
    x = "Distance between markets (km)", y = "Mean absolute gap (Rs/kg)"
  )
```

```{r}
#| label: lop-model
lm(mean_gap ~ distance_km, data = pair_gaps) |> summary()
```

Note `relationship = "many-to-many"` in the join. dplyr 1.1+ warns on unexpected
many-to-many joins because they are usually mistakes. Here it is intentional — we
*want* every market crossed with every market within a month — so we say so. Never
silence that warning by deleting it; declare intent instead.

## A composition-safe index {#sec-chained}

Back to Decision 4. The fix for an unbalanced panel is a **matched-sample chained
index**: compute the month-on-month price relative *within each market*, average
those relatives across only the markets present in both months, then chain.

```{r}
#| label: chained-index
wheat_chained <- wheat_market |>
  complete(market, date) |>                 # explicit gaps, so lag() is honest
  arrange(market, date) |>
  mutate(rel = price_pkr / lag(price_pkr), .by = market) |>
  summarise(
    gm        = exp(mean(log(rel), na.rm = TRUE)),   # geometric mean of relatives
    n_matched = sum(!is.na(rel)),
    .by = date
  ) |>
  arrange(date) |>
  mutate(
    gm    = replace_when(gm, is.na(gm) ~ 1),        # first month has no relative
    index = 100 * cumprod(gm)
  )
```

```{r}
#| label: index-comparison
naive_index <- wheat_national |>
  arrange(date) |>
  filter(!is.na(price_pkr)) |>
  mutate(index = 100 * price_pkr / first(price_pkr)) |>
  select(date, index) |>
  mutate(method = "Naive cross-market mean")

bind_rows(
  naive_index,
  wheat_chained |> select(date, index) |> mutate(method = "Matched-sample chained")
) |>
  ggplot(aes(date, index, colour = method)) +
  geom_line(linewidth = 0.7) +
  labs(
    title    = "Why composition matters",
    subtitle = "Two wheat flour indices from identical raw data",
    x = NULL, y = "Index (first month = 100)", colour = NULL
  )
```

If the two lines diverge materially, the naive average was measuring changes in who
reported, not changes in what things cost. Show this plot to anyone who asks why
your numbers differ from a published series.

## Indexing each market to its own start

Useful for a different question: which markets have seen the steepest increases?

```{r}
#| label: market-index
wheat_market_index <- wheat_market |>
  filter(!is.na(price_pkr)) |>
  arrange(market, date) |>
  mutate(index = 100 * price_pkr / first(price_pkr), .by = market)
```

Because `mutate()` with `.by` does not reorder rows, the `arrange()` beforehand is
load-bearing. `first()` returns the first row *as the data is currently sorted* —
this is a classic silent error.

# Visualisation

The plots below share a deliberate structure: one idea per chart, an informative
subtitle carrying the method, and a source caption. In official publications the
caption is not decoration — it is provenance.

```{r}
#| label: plot-staples
top_commodities <- basket |>
  count(commodity, sort = TRUE) |>
  slice_head(n = 4) |>
  pull(commodity)

p_staples <- national_monthly |>
  filter(commodity %in% top_commodities) |>
  ggplot(aes(date, price_pkr, colour = commodity)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = label_number(prefix = "Rs ")) +
  labs(
    title    = "Nominal staple food prices, Pakistan",
    subtitle = "Retail, actual observations, mean across reporting markets",
    x = NULL, y = "Rs per kg / litre", colour = NULL,
    caption  = "Source: WFP VAM Food Prices Database via HDX"
  )

p_staples
```

```{r}
#| label: plot-facets
p_markets <- wheat_market |>
  ggplot(aes(date, price_pkr)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ fct_reorder(market, price_pkr, .fun = max, .na_rm = TRUE)) +
  scale_y_continuous(labels = label_number(prefix = "Rs ")) +
  labs(title = "Wheat flour price by market",
       subtitle = "Panels ordered by peak price",
       x = NULL, y = NULL,
       caption = "Source: WFP VAM via HDX")

p_markets
```

Ordering facets by a meaningful statistic instead of alphabetically turns a wall
of panels into an argument.

```{r}
#| label: plot-index
p_index <- wheat_market_index |>
  ggplot(aes(date, index, group = market)) +
  geom_line(linewidth = 0.4, colour = "grey65") +
  geom_line(
    data = wheat_chained, aes(date, index, group = 1),
    linewidth = 1, colour = "firebrick"
  ) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40") +
  labs(
    title    = "Each market indexed to its own first observation",
    subtitle = "Grey = individual markets; red = matched-sample chained national index",
    x = NULL, y = "Index (= 100)",
    caption  = "Source: WFP VAM via HDX"
  )

p_index
```

```{r}
#| label: plot-yoy
p_yoy <- national_monthly |>
  filter(commodity %in% top_commodities, !is.na(yoy)) |>
  ggplot(aes(date, yoy, colour = commodity)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Year-on-year food price inflation", x = NULL, y = NULL, colour = NULL,
       caption = "Source: WFP VAM via HDX")

p_yoy
```

```{r}
#| label: plot-spread
p_spread <- dispersion |>
  ggplot(aes(date)) +
  geom_ribbon(aes(ymin = cheapest, ymax = dearest), alpha = 0.25, fill = "steelblue") +
  geom_line(aes(y = (cheapest + dearest) / 2), linewidth = 0.6) +
  scale_y_continuous(labels = label_number(prefix = "Rs ")) +
  labs(
    title    = "Cheapest and dearest market for wheat flour",
    subtitle = "Months with at least five reporting markets",
    x = NULL, y = "Rs per kg",
    caption  = "Source: WFP VAM via HDX"
  )

p_spread
```

A ribbon between the min and max carries more information than the spread alone —
the reader sees both the level and the dispersion.

```{r}
#| label: plot-heat
p_heat <- province_year |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)),
         !is.na(province)) |>
  ggplot(aes(year, fct_reorder(province, price_pkr), fill = price_pkr)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_viridis_c(option = "magma", direction = -1,
                       labels = label_number(prefix = "Rs ")) +
  scale_x_continuous(expand = expansion(0)) +
  labs(title = "Wheat flour by province and year", x = NULL, y = NULL, fill = NULL,
       caption = "Source: WFP VAM via HDX")

p_heat
```

# Mapping Pakistan, including Jammu & Kashmir {#sec-map}

This section deals with the part that most tutorials get wrong: a map of Pakistan
that omits Azad Jammu & Kashmir, Gilgit-Baltistan and the wider Jammu & Kashmir
region is not an acceptable output for a Pakistani institution, and a map that
shades a disputed territory as though it had data is not acceptable for anyone.

The approach: draw the full territory, show the disputed region with a distinct
"no data" treatment, and attach a disclaimer.

## Inspect before you join

Natural Earth uses its own spellings, and they change between releases. Always
look first.

```{r}
#| label: ne-inspect
pak_states <- ne_states(country = "Pakistan", returnclass = "sf") |> st_make_valid()
ind_states <- ne_states(country = "India",    returnclass = "sf") |> st_make_valid()

pak_states |>
  st_drop_geometry() |>
  distinct(name, type_en) |>
  arrange(name) |>
  print(n = Inf)

ind_states |>
  st_drop_geometry() |>
  distinct(name) |>
  filter(str_detect(name, regex("kashmir|jammu|ladakh", ignore_case = TRUE)))
```

Depending on your Natural Earth vintage you will see the Pakistan-administered
units as `Azad Kashmir` and either `Northern Areas` or `Gilgit-Baltistan`, and the
India-administered unit as `Jammu and Kashmir` (possibly split into `Ladakh` in
newer releases). The lookup table we built in @sec-verbs already handles both
spellings — this is the payoff for centralising the crosswalk.

## Assemble the base map

```{r}
#| label: base-map
pak_admin <- pak_states |>
  mutate(
    province = recode_values(
      name,
      from = province_lookup$from,
      to   = province_lookup$to
    ),
    status = "Administered by Pakistan"
  ) |>
  select(province, status, geometry)

# Anything Natural Earth calls something our lookup does not know
pak_admin |> st_drop_geometry() |> filter(is.na(province))
```

```{r}
#| label: jk-polygon
jk_geometry <- ind_states |>
  filter(str_detect(name, regex("jammu|kashmir|ladakh", ignore_case = TRUE))) |>
  st_geometry() |>
  st_union()                                    # dissolve into one polygon

jammu_kashmir <- st_sf(
  province = "Jammu & Kashmir",
  status   = "Disputed territory — no reporting market",
  geometry = jk_geometry
)

base_map <- bind_rows(pak_admin, jammu_kashmir) |>
  st_as_sf()

st_crs(base_map)
base_map |> st_drop_geometry() |> count(status)
```

Two details:

- `st_union()` dissolves Jammu & Kashmir and Ladakh into a single region so the
  internal administrative line does not appear on a Pakistani map.
- `bind_rows()` on `sf` objects works, but the result can lose its `sf` class in
  some combinations — the trailing `st_as_sf()` is cheap insurance.

## Check the geography joins to the data

```{r}
#| label: geo-join-check
province_recent <- province_year |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE))) |>
  filter(year == max(year)) |>
  summarise(price_pkr = mean(price_pkr, na.rm = TRUE), .by = province)

# Provinces in the data with no polygon — a real error, must be empty
province_recent |>
  anti_join(st_drop_geometry(base_map), by = join_by(province))

# Polygons with no data — expected, and the point of this section
base_map |>
  st_drop_geometry() |>
  anti_join(province_recent, by = join_by(province))
```

The first anti-join must return zero rows. If it does not, your recode is
incomplete and part of your data is falling off the map — literally. The second
anti-join is *supposed* to return rows: Jammu & Kashmir, and any province where no
market reported this year. Those become the grey areas.

## The choropleth

```{r}
#| label: choropleth
map_data <- base_map |>
  left_join(province_recent, by = join_by(province))

disputed <- map_data |> filter(str_detect(status, "Disputed"))

# Lambert conformal conic, sensible for a country of Pakistan's shape and latitude
pak_crs <- "+proj=lcc +lat_1=28 +lat_2=37 +lat_0=30 +lon_0=70 +datum=WGS84 +units=m"

p_map <- ggplot(map_data) +
  geom_sf(aes(fill = price_pkr), colour = "white", linewidth = 0.3) +
  geom_sf(
    data = disputed,
    aes(linetype = status),
    fill = "grey92", colour = "grey30", linewidth = 0.5
  ) +
  geom_sf_label(
    data  = disputed,
    aes(label = str_wrap(province, 12)),
    size = 3, label.size = 0, fill = alpha("white", 0.7)
  ) +
  scale_fill_viridis_c(
    option = "rocket", direction = -1, na.value = "grey88",
    labels = label_number(prefix = "Rs "),
    name   = "Wheat flour\nRs per kg"
  ) +
  scale_linetype_manual(values = c("22"), name = NULL) +
  coord_sf(crs = pak_crs) +
  labs(
    title    = paste0("Retail wheat flour price by province, ", max(province_year$year)),
    subtitle = "Grey areas have no reporting market in the WFP sample",
    caption  = paste(
      "Source: WFP VAM Food Prices Database via HDX; boundaries from Natural Earth.",
      "The boundaries and names shown do not imply official endorsement or acceptance.",
      "The final status of Jammu & Kashmir has not been agreed by the parties.",
      sep = "\n"
    )
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.caption      = element_text(hjust = 0, size = 8, colour = "grey30"),
    plot.title        = element_text(face = "bold"),
    legend.position   = "right"
  )

p_map
```

::: {.callout-important}
## On the disclaimer

The wording above follows the standard UN cartographic convention. Institutions in
Pakistan generally have their own approved formulation, and official publications
may additionally be required to show the Line of Control as a non-final boundary.
Check your organisation's map policy before publishing; the technical mechanism is
the same whichever wording you are required to use.

To add the Line of Control explicitly, Natural Earth ships a disputed-boundaries
line layer:

```{r}
#| label: loc-line
disputed_lines <- ne_download(
  scale       = 10,
  type        = "admin_0_boundary_lines_disputed_areas",
  category    = "cultural",
  returnclass = "sf"
)

p_map + geom_sf(data = disputed_lines, colour = "grey20",
                linewidth = 0.4, linetype = "13", inherit.aes = FALSE)
```
:::

## Market point map

```{r}
#| label: point-map
library(ggrepel)

market_recent <- wheat_market |>
  filter(date == max(date)) |>
  select(market, price_pkr)

markets_plot <- markets_sf |>
  left_join(market_recent, by = join_by(market)) |>
  filter(!is.na(price_pkr))

p_points <- ggplot() +
  geom_sf(data = base_map, fill = "grey96", colour = "white", linewidth = 0.3) +
  geom_sf(data = disputed, fill = "grey92", colour = "grey30",
          linewidth = 0.4, linetype = "22") +
  geom_sf(data = markets_plot, aes(size = price_pkr, colour = price_pkr)) +
  geom_text_repel(
    data = markets_plot,
    aes(label = market, geometry = geometry),
    stat = "sf_coordinates", size = 3, min.segment.length = 0,
    segment.colour = "grey50", max.overlaps = Inf
  ) +
  scale_colour_viridis_c(option = "rocket", direction = -1,
                         labels = label_number(prefix = "Rs ")) +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  coord_sf(crs = pak_crs) +
  labs(
    title    = "Reporting markets and latest wheat flour price",
    colour   = "Rs per kg",
    caption  = "Source: WFP VAM via HDX. Boundaries: Natural Earth.\nDisputed area shown for completeness; no data available."
  ) +
  theme_void(base_size = 12) +
  theme(plot.caption = element_text(hjust = 0, size = 8, colour = "grey30"))

p_points
```

`geom_text_repel(stat = "sf_coordinates")` is the correct way to label an `sf`
layer — it extracts coordinates from the geometry column and then repels, which
`geom_sf_text()` cannot do.

## Small multiples over time

```{r}
#| label: map-facets
snapshot_years <- c(2010, 2015, 2020, max(province_year$year))

map_panel <- province_year |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)),
         year %in% snapshot_years) |>
  summarise(price_pkr = mean(price_pkr, na.rm = TRUE), .by = c(province, year))

# Cross every polygon with every snapshot year, then join — so that provinces
# with no data still appear (grey) in every panel rather than vanishing.
p_map_facets <- base_map |>
  cross_join(tibble(year = snapshot_years)) |>
  left_join(map_panel, by = join_by(province, year)) |>
  st_as_sf() |>
  ggplot() +
  geom_sf(aes(fill = price_pkr), colour = "white", linewidth = 0.2) +
  facet_wrap(~ year) +
  scale_fill_viridis_c(option = "rocket", direction = -1, na.value = "grey88",
                       labels = label_number(prefix = "Rs ")) +
  coord_sf(crs = pak_crs) +
  labs(title = "Wheat flour price by province over time", fill = "Rs per kg",
       caption = "Grey = no reporting market or disputed territory.\nSource: WFP VAM via HDX.") +
  theme_void(base_size = 11) +
  theme(plot.caption = element_text(hjust = 0, size = 8, colour = "grey30"))

p_map_facets
```

`cross_join()` is the explicit way to build a complete polygon-year grid.
Left-joining onto that grid guarantees every province appears in every panel, which
is what makes the small multiples comparable. Joining the other way round — data
onto polygons — makes provinces silently disappear from panels where they have no
observation, and the reader has no way to tell absence of data from absence of
territory.

# Export and reproducibility

```{r}
#| label: export
out_fig  <- here("outputs", "figures")
out_data <- here("outputs", "data")
walk(c(out_fig, out_data), \(d) dir.create(d, recursive = TRUE, showWarnings = FALSE))

plots <- lst(
  p_staples, p_markets, p_index, p_yoy, p_spread, p_heat,
  p_map, p_points, p_map_facets
)

iwalk(plots, \(p, nm) {
  ggsave(
    filename = file.path(out_fig, paste0(nm, ".png")),
    plot     = p,
    width    = 9, height = 6, dpi = 300, bg = "white"
  )
})

write_csv(national_monthly, file.path(out_data, "national_monthly.csv"))
write_csv(province_year,    file.path(out_data, "province_year.csv"))
write_csv(wheat_chained,    file.path(out_data, "wheat_flour_index.csv"))
```

Three improvements over the obvious version:

- `lst()` auto-names the list from the object names, so `iwalk()` gets names for
  free. No parallel vector of filenames to fall out of sync.
- `bg = "white"` — `theme_minimal()` has a transparent background, and a
  transparent PNG dropped into a Word report renders on a grey or black panel.
- Figures and data go to separate folders, and `outputs/` should be in
  `.gitignore` — commit the code that makes them, not the artefacts.

```{r}
#| label: session
sessionInfo()
```

For a workshop, initialise `renv` on day one so every participant is on the same
package versions:

```{r}
#| label: renv
renv::init()
renv::snapshot()
```

# Common errors and what they mean

| Symptom | Likely cause | Fix |
|---|---|---|
| `could not find function "filter_out"` | dplyr < 1.2.0 | `install.packages("dplyr")` |
| Every value in a recoded column is `NA` | `recode_values()` on an incomplete mapping | Add cases, or use `replace_values()` |
| Row count grew after a join | Duplicate keys on one side | Check the grain with `count(key) \|> filter(n > 1)` |
| `Detected an unexpected many-to-many relationship` | Genuine duplicate keys, or an intended cross | Fix the grain, or declare `relationship = "many-to-many"` |
| Rows with `NA` disappeared after `filter()` | `filter()` treats `NA` as `FALSE` | Use `filter_out()` |
| `ymd()` returns all `NA` | Wrong date order in the source | Try `dmy()` / `mdy()`, or `parse_date()` |
| Provinces missing from the map | Recode mismatch between data and polygons | Run both `anti_join()` checks |
| `ne_states()` errors on scale 10 | `rnaturalearthhires` not installed | Install from the ropensci r-universe |
| Nonsensical year-on-year figures | `lag()` across reporting gaps | `complete()` the panel first |
| PNG has a grey background in Word | Transparent theme background | `ggsave(..., bg = "white")` |

# Exercises

**1. Basket construction.** Rewrite the `basket` filter to also include eggs, which
are priced per dozen. You will need a third branch in `when_any()` and a
`unit_to_dozen()` helper. How does adding eggs change `p_staples`?

**2. Wholesale margins.** Build a market-month table of the retail-to-wholesale
ratio for wheat. Which markets have the widest margins, and is the margin stable
over time?

**3. Sensitivity to Decision 1.** Re-run `national_monthly` keeping the `aggregate`
rows. Plot both versions on one chart. How much does the choice matter, and would
it change any conclusion you would draw?

**4. Provincial inflation.** Compute year-on-year inflation separately by province
using the matched-sample chained method from @sec-chained. Which province has had
the most volatile food prices?

**5. Winsorising sensitivity.** Change the winsorising threshold from the 99.9th to
the 99th percentile. Which results move? Which do not? Write two sentences you
would be willing to defend in a review meeting.

**6. Affordability by province.** Extend the affordability analysis to produce a
provincial choropleth of kilograms-per-daily-wage. Keep the Jammu & Kashmir
treatment from @sec-map.

**7. Districts.** The file has an `admin2` column. Build a district-level map for
one province. You will need district polygons — GADM via the `geodata` package, or
official boundaries from the relevant provincial bureau. Document which boundary
vintage you used and why.

**8. Seasonality by commodity.** Fit the seasonality model for each of the top four
commodities using `nest()` and `map()`, extract the month effects with
`marginaleffects`, and present them as a faceted chart.

# Where this goes next

Three natural extensions, roughly in order of effort:

- **A parameterised report.** Make `commodity` and `province` Quarto parameters and
  render one briefing note per province with `quarto::quarto_render()` in a loop.
- **A `targets` pipeline.** The cache-download-clean-analyse chain here is exactly
  what `targets` is for. The download becomes a tracked target with an explicit
  vintage, and re-running only recomputes what changed.
- **Joining to other sources.** PSLM and MICS give you household consumption
  shares; combine them with these prices to build a district cost-of-diet or a
  food poverty line. That is where price data starts answering policy questions
  rather than describing markets.
