---
title: "Reading, Cleaning and Mapping WFP Food Prices"
subtitle: "A step-by-step walkthrough of every command in the pipeline"
author: "Prof. Dr. Zahid Asghar"
institute: "School of Economics, QAU · d4d"
date: today
format:
  revealjs:
    theme: [default, d4d-navy-gold.scss]
    slide-number: true
    footer: "d4d · WFP Food Prices Pipeline"
    transition: fade
    code-line-numbers: true
    highlight-style: github
    scrollable: true
    chalkboard: false
execute:
  eval: false
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    comment: "#>"
---

## Why this session

- Two CSVs from HDX → **one clean pipeline** → national tables, inflation,
  affordability, market integration, and a map that includes the full
  territory of Pakistan.
- Every command you will run today is in this deck, in the order it runs.
- Nothing is hidden in a function you haven't seen — the two custom helpers
  (`unit_to_kg()`, `unit_to_litre()`) are shown and tested in full.
- **No R interpreter was used to write this deck.** Run each chunk yourself,
  in order, and check the audit line before moving to the next slide.

## How the pipeline is organised

::: {.incremental}
- **Part A** — read the two files, separately and simply
- **Part B1** — four analytical decisions that determine the headline numbers
- **Part B2–B3** — basket construction, province crosswalk, analysis frames
- **Part B4–B6** — inflation index, market integration, affordability
- **Part B7–B8** — the map, including Jammu & Kashmir, and the full plot set
- **Part B9** — export
:::

## Before you start

```{r}
#| label: setup
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)
library(scales)
library(here)
library(ggrepel)

theme_set(theme_minimal(base_size = 12))
```

Six packages beyond base tidyverse: `sf` for spatial data, `rnaturalearth` +
`rnaturalearthhires` for boundary polygons, `scales` for axis labels, `here`
for project-relative paths, `ggrepel` for map labels that don't overlap.

## Fail loudly, early

```{r}
#| label: version-check
if (packageVersion("dplyr") < "1.2.0") {
  stop("Needs dplyr >= 1.2.0; you have ", packageVersion("dplyr"))
}
```

- `filter_out()`, `when_any()`, `when_all()`, `recode_values()`,
  `replace_values()`, `replace_when()` are all **dplyr 1.2.0** verbs.
- Check the version **before** anything else fails with a confusing
  "could not find function" fifty lines later.

::: {.callout-note}
Two downloaded CSVs are assumed to already sit in `data/raw/`. Get them by
hand from the HDX dataset page before continuing.
:::

# Part A — Reading the Two Files {.section-divider background-color="#0f2440"}

## What Part A produces

Two clean, **separate** tibbles:

- `prices` — one row per market–commodity–month
- `markets` — one row per market (id, name, admin units, coordinates)

No custom functions, no API calls. If you only ever need prices, you can
stop after **A1** and skip **A2** entirely.

## A1 · See what you actually downloaded

```{r}
#| label: list-files
raw_dir <- here("data", "raw")

list.files(raw_dir, pattern = "\\.csv$")
```

Check the real filenames before typing them from memory in the next chunk.
This one line has saved more workshop time than any other in this deck.

## A1 · Read the prices file as text

```{r}
#| label: read-prices
prices_raw <- read_csv(
  file.path(raw_dir, "wfp_food_prices_pak.csv"),
  col_types = cols(.default = col_character())
)
```

Everything comes in as **character**, deliberately. Row 1 of every HDX file
is a HXL tag row (`#date`, `#adm1+name`, …), and letting `read_csv()` guess
column types on that row produces unpredictable results.

## A1 · Look at the row you're about to delete

```{r}
#| label: view-hxl
prices_raw |> slice(1)
```

Always **look** before you drop. This is the HXL tag row — a schema
annotation, not an observation. If this doesn't look like `#date`,
`#adm1+name`, … stop and check the file.

## A1 · Remove it

```{r}
#| label: drop-hxl
prices_raw <- prices_raw |> slice(-1)

glimpse(prices_raw)
```

`slice(-1)` removes row 1 positionally — the one rule that works whether or
not the file has a `date` column (the markets file, coming up, doesn't).

## A1 · Convert types and derive date parts

```{r}
#| label: clean-prices
prices <- prices_raw |>
  mutate(
    date      = ymd(date),
    price     = as.numeric(price),
    usdprice  = as.numeric(usdprice),
    latitude  = as.numeric(latitude),
    longitude = as.numeric(longitude),
    across(where(is.character), str_squish),
    year      = year(date),
    month_num = month(date),
    month_lab = month(date, label = TRUE, abbr = TRUE)
  )
```

`str_squish()` across every remaining character column strips stray
whitespace — a common HDX artefact that breaks exact-string matching later
without ever showing up in `glimpse()`.

## A1 · Audit the conversion

```{r}
#| label: audit-conversion
prices |> summarise(bad_date = sum(is.na(date)), bad_price = sum(is.na(price)))
```

Must be **0**, or close to it. Any coercion failure — a malformed date, a
non-numeric price — turns silently into `NA` at the `mutate()` step. This is
the only place you'll catch it.

## A1 · Get to know the columns

```{r}
#| label: count-checks
prices |> count(pricetype, priceflag, sort = TRUE)
prices |> count(unit, sort = TRUE)
prices |> count(admin1, sort = TRUE)
```

Three `count()` calls, three things you now know:

- Retail vs Wholesale, actual vs aggregate — **and their relative sizes**
- Every unit string you'll need to normalise later
- Every province spelling you'll need in the crosswalk

## A2 · Read the markets file — same recipe

```{r}
#| label: read-markets
markets_raw <- read_csv(
  file.path(raw_dir, "wfp_markets_pak.csv"),
  col_types = cols(.default = col_character())
)

markets_raw |> slice(1)          # look before deleting
markets_raw <- markets_raw |> slice(-1)

glimpse(markets_raw)
```

Identical pattern to A1: text first, view the HXL row, then drop it.
Consistency here is the point — one mental model for both files.

## A2 · Convert types

```{r}
#| label: clean-markets
markets <- markets_raw |>
  mutate(
    latitude  = as.numeric(latitude),
    longitude = as.numeric(longitude)
  )
```

Fewer columns need conversion than in the prices file — this is a static
reference table, not a time series.

## A2 · Confirm the grain

```{r}
#| label: markets-grain
markets |> count(market_id) |> filter(n > 1)
```

**Must return zero rows.** `markets` is meant to be one row per market. Any
join against this table — and you will join against it — silently
multiplies your data if this check fails. Fix duplicates before proceeding.

## A2 · What the markets file actually adds

```{r}
#| label: markets-no-data
markets_no_data <- markets |>
  anti_join(prices_raw, by = join_by(market_id))

markets_no_data |> select(market_id, market, admin1, admin2)
```

The prices file already carries coordinates and province names on every
row. What only the markets file can tell you: **which markets in the
sampling frame never reported a single price.**

## A2 · Quantify the coverage gap

```{r}
#| label: coverage-numbers
n_distinct(prices_raw$market_id)   # markets that reported at least once
nrow(markets_raw)                  # markets in the full sampling frame
nrow(markets_no_data)              # the gap between the two
```

::: {.callout-important}
This gap is a genuine finding — state it in any write-up: *"WFP maintains N
reporting markets for Pakistan; only M of them had a price observation in
the study period."*
:::

# Part B — The Analysis Pipeline {.section-divider background-color="#0f2440"}

## Four decisions before any average

::: {.incremental}
1. **actual vs aggregate** — keep only observed prices
2. **retail vs wholesale** — never pool the two
3. **units** — `KG`, `20 KG`, `500 G`, `L` are not comparable raw
4. **panel balance** — who reports, and when, changes the "national" mean
:::

Every one of these is invisible in `glimpse()`. Get any one wrong and the
headline number is still a number — just the wrong one.

## B1.1 · Actual, not aggregate

```{r}
#| label: decision-1
prices_actual <- prices |> filter_out(priceflag != "actual")
```

`aggregate` rows are WFP-computed roll-ups, not observed prices — keeping
them double-counts. `filter_out()` drops rows matching the condition **and**
keeps rows where `priceflag` is missing, unlike a plain `filter(!(...))`
which would silently lose them.

## B1.2 · Retail only

```{r}
#| label: decision-2
retail <- prices_actual |> filter(pricetype == "Retail")
```

The most commonly missed decision. Pool Retail and Wholesale and your
"national price" moves whenever the **mix** of reporting types changes —
even if no actual price moved anywhere.

## B1.3 · Helper 1 — mass units to kilograms

```{r}
#| label: unit-to-kg
unit_to_kg <- function(unit) {
  qty  <- coalesce(parse_number(unit), 1)
  mult <- case_when(
    str_detect(unit, regex("\\bMT\\b", ignore_case = TRUE)) ~ 1000,
    str_detect(unit, regex("\\bKG\\b", ignore_case = TRUE)) ~ 1,
    str_detect(unit, regex("\\bG\\b",  ignore_case = TRUE)) ~ 1 / 1000,
    .default = NA_real_
  )
  qty * mult
}
```

`parse_number("20 KG")` gives `20`; `parse_number("KG")` gives `NA`, so
`coalesce(..., 1)` supplies the implied quantity of one.

## B1.3 · Helper 2 — volume units to litres

```{r}
#| label: unit-to-litre
unit_to_litre <- function(unit) {
  qty  <- coalesce(parse_number(unit), 1)
  mult <- case_when(
    str_detect(unit, regex("\\bML\\b", ignore_case = TRUE)) ~ 1 / 1000,
    str_detect(unit, regex("\\bL\\b",  ignore_case = TRUE)) ~ 1,
    .default = NA_real_
  )
  qty * mult
}
```

Order matters in `case_when()` — first match wins. `MT` before `KG`,
`ML` before `L`. The `\\b` word boundaries stop `G` matching inside `KG`.

## B1.3 · Test before you trust

```{r}
#| label: test-units
test_units <- c("KG", "20 KG", "500 G", "MT", "L", "1.5 L", "Unit", "Day", "Dozen")

tibble(
  unit  = test_units,
  kg    = unit_to_kg(test_units),
  litre = unit_to_litre(test_units)
)
```

`unit_to_kg("20 KG")` should read `20`, not `1`. `unit_to_litre("KG")`
should read `NA`, not `0`. Run this **before** applying the function to
13,000 rows.

## B1.3 · Apply, and audit what's left {.smaller}

```{r}
#| label: apply-units
retail <- retail |>
  mutate(
    kg_equiv     = unit_to_kg(unit),
    l_equiv      = unit_to_litre(unit),
    price_per_kg = price / kg_equiv,
    price_per_l  = price / l_equiv,
    price_std    = coalesce(price_per_kg, price_per_l),
    unit_std     = case_when(
      !is.na(price_per_kg) ~ "per kg",
      !is.na(price_per_l)  ~ "per litre",
      .default = NA_character_
    )
  )

# nothing food-and-mass-based should appear here
retail |> filter(is.na(unit_std)) |> count(category, commodity, unit, sort = TRUE)
```

Expect to see wages (`Day`), fuel, eggs (`Dozen`) in the unclassified list.
If a cereal or a pulse shows up, extend the function — don't drop the rows.

## B1.4 · Look before you average

```{r}
#| label: decision-4
retail |>
  filter(!is.na(price_std)) |>
  summarise(n_obs = n(), n_months = n_distinct(date),
            first_obs = min(date), last_obs = max(date),
            .by = c(commodity, market))
```

Markets enter and leave the sample. This table is the first warning sign —
we come back to fix the consequence in **B4**.

## B2 · The basket — `when_any()` / `when_all()`

```{r}
#| label: basket
basket <- retail |>
  filter_out(category == "non-food") |>
  filter(
    when_any(
      when_all(unit_std == "per kg",    category != "milk and dairy"),
      when_all(unit_std == "per litre",
               str_detect(commodity, regex("oil|ghee", ignore_case = TRUE)))
    )
  )
```

Read it aloud: *keep a row if (it's per-kg and not dairy) OR (it's per-litre
and it's an oil/ghee).* The code has the same shape as the sentence —
that's the entire design intent of these two verbs.

## B2 · One province lookup, used three times

```{r}
#| label: province-lookup
province_lookup <- tribble(
  ~from,                                                              ~to,
  c("Punjab"),                                                        "Punjab",
  c("Sindh"),                                                         "Sindh",
  c("Khyber Pakhtunkhwa", "N.W.F.P.", "NWFP",
    "Federally Administered Tribal Areas", "F.A.T.A."),               "KP",
  c("Balochistan", "Baluchistan"),                                    "Balochistan",
  c("Islamabad", "F.C.T.", "Islamabad Capital Territory"),            "Islamabad",
  c("Azad Kashmir", "Azad Jammu and Kashmir", "AJK"),                 "Azad Jammu & Kashmir",
  c("Gilgit-Baltistan", "Northern Areas"),                            "Gilgit-Baltistan"
)
```

A cell holding a **vector** means "any of these map to this." One table,
used for the price data, the markets file, **and** the map polygons later.
This is what keeps three separate name-cleaning jobs in sync.

## B2 · Apply it, and audit

```{r}
#| label: apply-province
basket <- basket |>
  mutate(province = recode_values(admin1,
                                  from = province_lookup$from,
                                  to   = province_lookup$to))

# must be empty; if not, extend province_lookup
basket |> filter(is.na(province), !is.na(admin1)) |> count(admin1, sort = TRUE)
```

`recode_values()` is a **total** mapping — anything not listed becomes
`NA`. That's why the audit line exists: an incomplete lookup fails silently
without it.

## B2 · Winsorise with `replace_when()`

```{r}
#| label: winsorise
basket <- basket |>
  mutate(
    price_w = replace_when(
      price_std,
      price_std > quantile(price_std, 0.999, na.rm = TRUE)
      ~ quantile(price_std, 0.999, na.rm = TRUE)
    ),
    .by = c(commodity, unit_std)
  )

basket |>
  filter(price_std != price_w) |>
  select(date, market, commodity, price_std, price_w) |>
  arrange(desc(price_std))
```

`replace_when()` is **partial** — everything not matched passes through
unchanged. Always print what changed. Winsorising silently is not
defensible; winsorising visibly is.

## B3 · `market_meta` — authoritative, not reconstructed

```{r}
#| label: market-meta
market_meta <- markets |>
  mutate(province = recode_values(admin1,
                                  from = province_lookup$from,
                                  to   = province_lookup$to)) |>
  select(market_id, market, admin1, admin2, province, latitude, longitude) |>
  mutate(reporting = !market_id %in% markets_no_data$market_id)

market_meta |> filter(is.na(province), !is.na(admin1)) |> count(admin1, sort = TRUE)
market_meta |> count(reporting)
```

Built from the **markets file**, not from `distinct()` on `basket`. A
`market_meta` built from `basket` silently loses every market with zero
observations. This version carries a `reporting` flag instead — nothing is
lost, and the map later can show it.

## B3 · National monthly series

```{r}
#| label: national-monthly
national_monthly <- basket |>
  summarise(price_pkr = mean(price_w,  na.rm = TRUE),
            price_usd = mean(usdprice, na.rm = TRUE),
            n_markets = n_distinct(market),
            .by = c(date, commodity, unit_std)) |>
  complete(commodity, date) |>                  # honest gaps before any lag()
  arrange(commodity, date) |>
  mutate(mom = price_pkr / lag(price_pkr)     - 1,
         yoy = price_pkr / lag(price_pkr, 12) - 1,
         .by = commodity)
```

`lag()` counts **rows**, not months. Without `complete()` first, a
reporting gap fabricates a year-on-year figure — twelve rows back is not
the same as twelve months back once a market drops out for a while.

## B3 · Provincial series and the reference commodity

```{r}
#| label: province-year
province_year <- basket |>
  summarise(price_pkr = mean(price_w, na.rm = TRUE),
            n_markets = n_distinct(market),
            .by = c(province, commodity, year))

# pick the reference commodity FROM THE DATA, never from memory
basket |> count(commodity, sort = TRUE) |> print(n = Inf)
```

Copy the exact commodity string from that `count()` output. Never hard-code
`"Wheat flour"` from memory — vintages and spellings change.

## B3 · The wheat frames

```{r}
#| label: wheat-frames
wheat <- basket |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)))

wheat_market <- wheat |>
  summarise(price_pkr = mean(price_w, na.rm = TRUE),
            .by = c(date, market, province))

wheat_national <- national_monthly |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)))
```

Three grains, named explicitly: `wheat` is one row per observation,
`wheat_market` one row per market-month, `wheat_national` one row per month.

## B4 · Why a plain average is composition-varying

If an expensive market starts reporting mid-series, the raw cross-market
mean jumps — with no price change anywhere. Fix: a **matched-sample
chained index**.

## B4 · Build the chained index

```{r}
#| label: chained-index
wheat_chained <- wheat_market |>
  complete(market, date) |>
  arrange(market, date) |>
  mutate(rel = price_pkr / lag(price_pkr), .by = market) |>
  summarise(gm        = exp(mean(log(rel), na.rm = TRUE)),
            n_matched = sum(!is.na(rel)),
            .by = date) |>
  arrange(date) |>
  mutate(gm    = replace_when(gm, is.na(gm) ~ 1),
         index = 100 * cumprod(gm))
```

For each market compute its own month-on-month relative, take the
**geometric mean of relatives** across only markets present in both months,
then chain with `cumprod()`. This is composition-safe by construction.

## B4 · Per-market index (a different question)

```{r}
#| label: market-index
wheat_market_index <- wheat_market |>
  filter(!is.na(price_pkr)) |>
  arrange(market, date) |>                      # load-bearing
  mutate(index = 100 * price_pkr / first(price_pkr), .by = market)
```

`.by` does not reorder rows. `first()` returns the first row **as currently
sorted** — the `arrange()` before it is not decorative, it's required.

## B5 · Dispersion across markets

```{r}
#| label: dispersion
dispersion <- wheat_market |>
  filter(!is.na(price_pkr)) |>
  summarise(n_markets = n(),
            cheapest  = min(price_pkr),
            dearest   = max(price_pkr),
            spread    = dearest - cheapest,
            cv        = sd(price_pkr) / mean(price_pkr),
            .by = date) |>
  filter(n_markets >= 5)
```

`min()`/`max()` widen **mechanically** as the sample grows. Comparing a
3-market spread to a 20-market spread tells you about coverage, not
integration — hence the `n_markets >= 5` floor.

## B5 · Distances between markets

```{r}
#| label: distances
markets_sf <- market_meta |>
  filter_out(is.na(latitude) | is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

dist_km <- st_distance(markets_sf) |>
  units::set_units("km") |>
  units::drop_units()
dimnames(dist_km) <- list(markets_sf$market, markets_sf$market)

pair_distance <- dist_km |>
  as_tibble(rownames = "market_a") |>
  pivot_longer(-market_a, names_to = "market_b", values_to = "distance_km") |>
  filter(market_a < market_b)
```

`market_a < market_b` (string comparison) keeps one row per pair instead of
two mirrored rows — a small trick worth remembering for any symmetric
pairwise table.

## B5 · Law of one price {.smaller}

```{r}
#| label: pair-gaps
pair_gaps <- wheat_market |>
  select(date, market, price_pkr) |>
  inner_join(wheat_market |> select(date, market, price_pkr),
             by           = join_by(date),
             relationship = "many-to-many",     # intentional cross, declared
             suffix       = c("_a", "_b")) |>
  filter(market_a < market_b) |>
  mutate(gap = abs(price_pkr_a - price_pkr_b)) |>
  summarise(mean_gap = mean(gap, na.rm = TRUE), n_months = n(),
            .by = c(market_a, market_b)) |>
  filter(n_months >= 24) |>
  inner_join(pair_distance, by = join_by(market_a, market_b))

summary(lm(mean_gap ~ distance_km, data = pair_gaps))
```

`relationship = "many-to-many"` is declared, not suppressed — dplyr warns
on unexpected many-to-many joins because they're usually bugs. Here it's
the deliberate point: cross every market against every other market.

## B6 · Affordability

```{r}
#| label: affordability
wages <- prices_actual |>
  filter(str_detect(commodity, regex("wage", ignore_case = TRUE))) |>
  summarise(wage_pkr = mean(price, na.rm = TRUE), .by = c(date, market))

affordability <- wheat_market |>
  inner_join(wages, by = join_by(date, market)) |>
  mutate(kg_per_day_wage = wage_pkr / price_pkr)

c(wheat_market = nrow(wheat_market), affordability = nrow(affordability))
```

The wage series lives inside `non-food` — exactly why it wasn't thrown away
in B2. `inner_join()` restricts you to months where **both** a price and a
wage were reported; the row-count comparison shows what that costs.

## B7 · The map, including Jammu & Kashmir

::: {.incremental}
- Boundaries omitting AJK/Gilgit-Baltistan are unacceptable from a Pakistani
  institution
- Shading a disputed territory as though it had data is unacceptable for
  anyone
- The approach: draw the full territory, show the disputed region as an
  explicit **no-data** layer, attach a disclaimer
:::

## B7 · Fetch and inspect boundaries

```{r}
#| label: fetch-boundaries
pak_states <- ne_states(country = "Pakistan", returnclass = "sf") |> st_make_valid()
ind_states <- ne_states(country = "India",    returnclass = "sf") |> st_make_valid()

# inspect before joining — Natural Earth spellings change between releases
pak_states |> st_drop_geometry() |> pull(name) |> unique() |> sort()
```

`st_drop_geometry()` returns a base data frame, not a tibble — `pull()`
sidesteps the `print(n=)` / `na.print` error that `distinct() |> print(n =
Inf)` would raise on it.

## B7 · Recode province names, audit

```{r}
#| label: pak-admin
pak_admin <- pak_states |>
  mutate(province = recode_values(name,
                                  from = province_lookup$from,
                                  to   = province_lookup$to),
         status   = "Administered by Pakistan") |>
  select(province, status, geometry)

# polygons our lookup does not recognise
pak_admin |> st_drop_geometry() |> filter(is.na(province))
```

Same `province_lookup` as B2 — the payoff for centralising it. Must be
empty; anything else is a spelling the lookup hasn't seen yet.

## B7 · Dissolve Jammu & Kashmir into one polygon

```{r}
#| label: jk-polygon
jk_geometry <- ind_states |>
  filter(str_detect(name, regex("jammu|kashmir|ladakh", ignore_case = TRUE))) |>
  st_geometry() |>
  st_union()

jammu_kashmir <- st_sf(
  province = "Jammu & Kashmir",
  status   = "Disputed territory — no reporting market",
  geometry = jk_geometry
)
```

`st_union()` dissolves the internal India-administered subdivisions so no
internal line appears on a Pakistani-facing map. `st_sf()` builds the
one-row spatial frame explicitly, sidestepping any class-dropping quirks in
`bind_rows()` on `sf` objects.

## B7 · Assemble and audit the base map

```{r}
#| label: base-map
base_map <- bind_rows(pak_admin, jammu_kashmir) |> st_as_sf()

province_recent <- province_year |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)),
         year == max(year)) |>
  summarise(price_pkr = mean(price_pkr, na.rm = TRUE), .by = province)

# AUDIT 1 — must be EMPTY: data with no polygon falls off the map
province_recent |> anti_join(st_drop_geometry(base_map), by = join_by(province))

# AUDIT 2 — expected to have rows: these become the grey "no data" areas
base_map |> st_drop_geometry() |> anti_join(province_recent, by = join_by(province))
```

Two audits, two different expected answers. The first must be zero rows —
your own province with no matching polygon is a real bug. The second is
supposed to return rows: Jammu & Kashmir, and any province with no market
this year.

## B8 · The core price plots

```{r}
#| label: plot-lines
top_commodities <- basket |> count(commodity, sort = TRUE) |>
  slice_head(n = 4) |> pull(commodity)

p_staples <- national_monthly |>
  filter(commodity %in% top_commodities) |>
  ggplot(aes(date, price_pkr, colour = commodity)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = label_number(prefix = "Rs ")) +
  labs(title = "Nominal staple food prices, Pakistan",
       subtitle = "Retail, actual observations, mean across reporting markets",
       x = NULL, y = "Rs per kg / litre", colour = NULL,
       caption = "Source: WFP VAM Food Prices Database via HDX")
```

Top-4 commodities picked **from the data**, not hard-coded — the same
discipline as picking `wheat` in B3.

## B8 · Facets ordered meaningfully

```{r}
#| label: plot-facets
p_markets <- wheat_market |>
  ggplot(aes(date, price_pkr)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ fct_reorder(market, price_pkr, .fun = max, .na_rm = TRUE)) +
  scale_y_continuous(labels = label_number(prefix = "Rs ")) +
  labs(title = "Wheat flour price by market", subtitle = "Panels ordered by peak price",
       x = NULL, y = NULL, caption = "Source: WFP VAM via HDX")
```

`fct_reorder()` by peak price turns a wall of alphabetised panels into an
argument — the reader sees the expensive markets first.

## B8 · Index: markets against the national line

```{r}
#| label: plot-index
p_index <- wheat_market_index |>
  ggplot(aes(date, index, group = market)) +
  geom_line(linewidth = 0.4, colour = "grey65") +
  geom_line(data = wheat_chained, aes(date, index, group = 1),
            linewidth = 1, colour = "firebrick") +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey40") +
  labs(title = "Each market indexed to its own first observation",
       subtitle = "Grey = markets; red = chained national index")
```

Plotting the composition-safe index **on top of** the raw market lines lets
the audience see immediately whether the national series is representative.

## B8 · Inflation and dispersion

```{r}
#| label: plot-yoy-spread
p_yoy <- national_monthly |>
  filter(commodity %in% top_commodities, !is.na(yoy)) |>
  ggplot(aes(date, yoy, colour = commodity)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Year-on-year food price inflation", colour = NULL)

p_spread <- dispersion |>
  ggplot(aes(date)) +
  geom_ribbon(aes(ymin = cheapest, ymax = dearest),
              alpha = 0.25, fill = "steelblue") +
  geom_line(aes(y = (cheapest + dearest) / 2), linewidth = 0.6) +
  labs(title = "Cheapest and dearest market for wheat flour")
```

The ribbon carries level **and** dispersion at once — more information than
a single spread line.

## B8 · Affordability plot

```{r}
#| label: plot-afford
p_afford <- affordability |>
  summarise(kg_per_day_wage = mean(kg_per_day_wage, na.rm = TRUE), .by = date) |>
  ggplot(aes(date, kg_per_day_wage)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, colour = "firebrick") +
  labs(title = "How much wheat flour does a day of unskilled labour buy?",
       y = "kg per day's wage")
```

Often more persuasive to a policy audience than any price series, because
it answers the question they actually have.

## B8 · Map setup — CRS, join, disclaimer

```{r}
#| label: map-setup
pak_crs <- "+proj=lcc +lat_1=28 +lat_2=37 +lat_0=30 +lon_0=70 +datum=WGS84 +units=m"

map_data <- base_map |> left_join(province_recent, by = join_by(province))
disputed <- map_data |> filter(str_detect(status, "Disputed"))

map_caption <- paste(
  "Source: WFP VAM Food Prices Database via HDX; boundaries from Natural Earth.",
  "The boundaries and names shown do not imply official endorsement or acceptance.",
  "The final status of Jammu & Kashmir has not been agreed by the parties.",
  sep = "\n"
)
```

Lambert conformal conic — sensible for a country of Pakistan's shape and
latitude. The disclaimer follows the UN cartographic convention; check your
institution's approved wording before publishing.

## B8 · The choropleth

```{r}
#| label: plot-choropleth
p_map <- ggplot(map_data) +
  geom_sf(aes(fill = price_pkr), colour = "white", linewidth = 0.3) +
  geom_sf(data = disputed, aes(linetype = status),
          fill = "grey92", colour = "grey30", linewidth = 0.5) +
  geom_sf_label(data = disputed, aes(label = str_wrap(province, 12)),
                size = 3, label.size = 0, fill = alpha("white", 0.7)) +
  scale_fill_viridis_c(option = "rocket", direction = -1, na.value = "grey88",
                       name = "Wheat flour\nRs per kg") +
  coord_sf(crs = pak_crs) +
  labs(title = "Retail wheat flour price by province",
       subtitle = "Grey areas have no reporting market in the WFP sample",
       caption = map_caption) +
  theme_void(base_size = 12)
```

The disputed layer gets its own `geom_sf()` call, its own linetype legend
key, and a label — it is visually distinct, not just differently coloured.

## B8 · Reporting vs non-reporting markets {.smaller}

```{r}
#| label: plot-points
market_recent <- wheat_market |> filter(date == max(date)) |> select(market, price_pkr)

markets_plot <- markets_sf |> left_join(market_recent, by = join_by(market))

p_points <- ggplot() +
  geom_sf(data = base_map, fill = "grey96", colour = "white", linewidth = 0.3) +
  geom_sf(data = disputed, fill = "grey92", colour = "grey30", linetype = "22") +
  geom_sf(data = filter(markets_plot, !reporting),
          shape = 21, colour = "grey60", fill = "white", size = 2) +
  geom_sf(data = filter(markets_plot, reporting, !is.na(price_pkr)),
          aes(size = price_pkr, colour = price_pkr)) +
  geom_text_repel(data = filter(markets_plot, reporting, !is.na(price_pkr)),
                  aes(label = market, geometry = geometry),
                  stat = "sf_coordinates", max.overlaps = Inf) +
  labs(title = "Reporting markets and latest wheat flour price",
       subtitle = "Hollow points = markets in the sampling frame with no observations")
```

This is where the `reporting` flag from **B3** pays off: hollow grey points
for markets that never reported, filled coloured points for the rest. The
map becomes a coverage diagnostic, not just a price map.

## B8 · Small multiples over time {.smaller}

```{r}
#| label: plot-facets-map
snapshot_years <- c(2010, 2015, 2020, max(province_year$year))

map_panel <- province_year |>
  filter(str_detect(commodity, regex("^wheat flour", ignore_case = TRUE)),
         year %in% snapshot_years) |>
  summarise(price_pkr = mean(price_pkr, na.rm = TRUE), .by = c(province, year))

p_map_facets <- base_map |>
  cross_join(tibble(year = snapshot_years)) |>   # complete polygon-year grid
  left_join(map_panel, by = join_by(province, year)) |>
  st_as_sf() |>
  ggplot() +
  geom_sf(aes(fill = price_pkr), colour = "white", linewidth = 0.2) +
  facet_wrap(~ year) +
  coord_sf(crs = pak_crs) +
  labs(title = "Wheat flour price by province over time", caption = map_caption)
```

`cross_join()` builds the complete polygon-year grid **first**. Join data
onto polygons the other way round and a province with no observation in a
given year simply vanishes from that panel — the reader can't tell absence
of data from absence of territory.

## B9 · Export, named without drift

```{r}
#| label: export
out_fig  <- here("outputs", "figures")
out_data <- here("outputs", "data")
walk(c(out_fig, out_data), \(d) dir.create(d, recursive = TRUE, showWarnings = FALSE))

plots <- lst(p_staples, p_markets, p_index, p_yoy, p_spread, p_afford,
             p_map, p_points, p_map_facets)

iwalk(plots, \(p, nm) {
  ggsave(file.path(out_fig, paste0(nm, ".png")), p,
         width = 9, height = 6, dpi = 300, bg = "white")
})
```

`lst()` auto-names the list from the object names, so `iwalk()` gets
filenames for free — no parallel vector to fall out of sync.
`bg = "white"` matters: `theme_minimal()` has a transparent background that
renders grey/black when a PNG lands inside Word.

## B9 · Write the data out

```{r}
#| label: write-csv
write_csv(national_monthly, file.path(out_data, "national_monthly.csv"))
write_csv(province_year,    file.path(out_data, "province_year.csv"))
write_csv(wheat_chained,    file.path(out_data, "wheat_flour_index.csv"))
write_csv(market_meta,      file.path(out_data, "market_meta.csv"))

sessionInfo()
```

`market_meta` is written out too — the reporting/non-reporting distinction
is a deliverable in its own right, not just an internal helper table.

# Recap {.section-divider background-color="#0f2440"}

## The one-slide summary {.smaller}

| Step | Command | What it protects against |
|---|---|---|
| Read | `slice(-1)` after character read | HXL row corrupting type guessing |
| Filter | `filter_out()` | `NA` silently dropped by plain negation |
| Combine conditions | `when_any()` / `when_all()` | Unreadable nested `&`/`\|` |
| Recode | `recode_values()` (total) vs `replace_values()` (partial) | Losing unmapped categories to `NA` |
| Edit values | `replace_when()` | Winsorising invisibly |
| Time series | `complete()` before `lag()` | Fabricated month-over-month change |
| Joins | `relationship =` declared | Silent row multiplication |
| Index | matched-sample chained | Composition-varying "average" |
| Map | dissolve + no-data layer + disclaimer | Omitting or misrepresenting territory |

## Questions

Run each chunk in order. If something errors, the audit line right before
it usually tells you which assumption broke.

**d4d · github.com/Zahedasghar/d4d**
