---
title: "Combining the WFP Prices and Markets Files"
subtitle: "A short workflow for two related HDX downloads"
author: "Prof. Dr. Zahid Asghar"
date: today
format:
  html:
    theme: cosmo
    toc: true
    code-copy: true
    df-print: paged
execute:
  eval: false
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    comment: "#>"
---

::: {.callout-note appearance="simple"}
This is a companion note to the main tutorial, for the specific case where you
download **two** files from the WFP Food Prices for Pakistan HDX page: the price
observations and the market/geodata reference table. It walks through the one
question that actually matters — *what does the second file give you that the
first doesn't?* — rather than assuming a join is automatically needed.
:::

# Why there are two files at all

The prices file is one row per market-commodity-month. The markets file is one
row per market — a static reference table: id, name, admin units, coordinates.
In principle the markets file is the authoritative source for geography, and the
prices file just repeats it on every row for convenience.

**Do not assume they agree, and do not assume you need both.** Check first.

# Step 1 — Read both files the same way

Both HDX files carry a HXL tag row (`#date`, `#adm1+name`, ...) as row 1. Read
everything as character, strip that row, and only convert types afterwards.

```{r}
#| label: read
library(tidyverse)
library(here)

raw_dir <- here("data", "raw")
list.files(raw_dir, pattern = "\\.csv$")   # confirm the actual filenames first

read_hdx <- function(path) {
  read_csv(path, col_types = cols(.default = col_character())) |>
    slice(-1) |>                                 # HXL row — always row 1, both files
    mutate(across(where(is.character), str_squish))
}

prices_raw  <- read_hdx(file.path(raw_dir, "wfp_food_prices_pak.csv"))
markets_raw <- read_hdx(file.path(raw_dir, "wfp_markets_pak.csv"))

glimpse(prices_raw)
glimpse(markets_raw)
```

::: {.callout-tip}
`slice(-1)` rather than `filter_out(str_starts(date, "#"))` — the markets file has
no `date` column, so positional removal is the one rule that works for either file.
:::

# Step 2 — Check the schema before writing a single `select()`

The single most common error here is guessing a column name (`market_name`) that
turns out not to exist. Always look first.

```{r}
#| label: schema
names(prices_raw)
names(markets_raw)
intersect(names(prices_raw), names(markets_raw))
```

In this dataset both files carry `market_id`, `market`, `admin1`, `admin2`,
`latitude`, `longitude`. That overlap is the second thing to notice: a naive join
on `market_id` will duplicate every one of those columns with `.x`/`.y` suffixes
unless you decide explicitly which side wins.

# Step 3 — Confirm the markets file is one row per market

Any join against `markets_raw` is only safe if this is true.

```{r}
#| label: key-check
markets_raw |> count(market_id) |> filter(n > 1)
```

Must return zero rows. If it doesn't, resolve the duplicates before going further
— do not join against a table with a broken grain.

# Step 4 — Do the two files actually agree?

Before deciding what to do with the overlap, check whether it's even a
disagreement worth resolving.

```{r}
#| label: reconcile
check <- prices_raw |>
  distinct(market_id, market, admin1, admin2, latitude, longitude) |>
  mutate(across(c(latitude, longitude), as.numeric)) |>
  inner_join(
    markets_raw |> mutate(across(c(latitude, longitude), as.numeric)),
    by     = join_by(market_id),
    suffix = c("_prices", "_markets")
  )

check |> filter(market_prices != market_markets)
check |> filter(admin1_prices != admin1_markets)
check |> filter(abs(latitude_prices - latitude_markets) > 0.01)
```

**Two possible outcomes, two different next steps:**

- **All three return zero rows** (the common case). The files fully agree. There
  is nothing to reconcile, and no metadata join is needed — the prices file is
  self-sufficient for geography. Skip to Step 5.
- **Any return rows.** Pick one file as authoritative (usually the markets file —
  it's the reference table) and drop the losing columns from the prices file
  before joining:

  ```{r}
  #| label: reconcile-fix
  #| eval: false
  prices <- prices_raw |>
    select(-market, -admin1, -admin2, -latitude, -longitude)

  combined <- prices |>
    left_join(
      markets_raw |> mutate(across(c(latitude, longitude), as.numeric)),
      by           = join_by(market_id),
      relationship = "many-to-one"
    )

  stopifnot(nrow(combined) == nrow(prices))   # row count must not change
  ```

# Step 5 — What the markets file actually gives you

If Step 4 came back clean, the markets file's real value isn't columns to add —
it's the markets that **never reported a price at all**. Only the reference table
knows that; a price table by definition has no rows for a market with no prices.

```{r}
#| label: coverage-gap
markets <- markets_raw |>
  mutate(across(c(latitude, longitude), as.numeric))

markets_no_data <- markets |>
  anti_join(prices_raw, by = join_by(market_id))

markets_no_data |> select(market_id, market, admin1, admin2)

n_distinct(prices_raw$market_id)   # markets that reported at least once
nrow(markets_raw)                  # markets in the full sampling frame
nrow(markets_no_data)              # the coverage gap
```

That gap is a finding, not a nuisance — state it: *"WFP maintains N reporting
markets for Pakistan; M of them had at least one price observation in the study
period."*

# Step 6 — Build one authoritative market table

```{r}
#| label: market-meta
prices <- prices_raw |>
  mutate(
    date = ymd(date),
    across(c(latitude, longitude, price, usdprice), as.numeric)
  )

market_meta <- prices |>
  distinct(market_id, market, admin1, admin2, latitude, longitude) |>
  mutate(reporting = TRUE) |>
  bind_rows(
    markets_no_data |> mutate(reporting = FALSE)
  )

market_meta |> count(reporting)

# Every market in the frame accounted for exactly once
stopifnot(nrow(market_meta) == n_distinct(prices_raw$market_id) + nrow(markets_no_data))
```

`reporting` is worth carrying all the way to the map: shade non-reporting markets
as hollow grey points on `p_points` instead of dropping them, and the figure
becomes a coverage diagnostic rather than just a price map.

# Decision summary

| Question | If yes | If no |
|---|---|---|
| Is `markets_raw` one row per `market_id`? | Proceed | Fix duplicates first — never join against a broken key |
| Do the files agree on shared columns (Step 4)? | Prices file is self-sufficient; skip the join | Pick one authoritative source and drop the loser's columns before joining |
| Do you need markets with zero observations? | Use `anti_join()` to isolate them; keep as a separate table | `prices` alone is enough |

# Exercise

Redo Step 4 for the `admin2` (district) column, which was not checked above.
Does agreement hold at the district level as well as the province level? If any
mismatches turn up, is the prices file or the markets file more likely to be
right, and how would you decide?
