---
title: "Group [Number] Project"
subtitle: "R & Positron Training Project"
author: "Group Member Names"
date: today
format:
  revealjs:
    theme: [default, custom.scss]

---

# Introduction

## Project Overview
- Brief description of what you'll demonstrate
- Dataset used
- Key objectives

## Group Members
- Member 1
- Member 2
- Member 3
- Member 4

---

# Data & Setup

## Loading Required Packages
```{r}
#| echo: true
#| message: false

library(tidyverse)
# Add other packages as needed
```

## Data Import
```{r}
#| echo: true
#| eval: false

# Load your dataset
data <- read_csv("your_data.csv")

# Preview the data
head(data)
```

---

# Analysis/Methodology

## Step 1: [First Task]
```{r}
#| echo: true

# Your code here
```

**Key Points:**
- What this code does
- Why it's important

## Step 2: [Second Task]
```{r}
#| echo: true

# Your code here
```

---

# Results

## Main Findings
```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Create visualization or output
```

## Interpretation
- Finding 1
- Finding 2
- Finding 3

---

# Positron Features

## What We Used
- Feature 1 and how it helped
- Feature 2 and its benefit
- Feature 3 demonstration

---

# Conclusion

## Key Takeaways
- Main lesson learned
- Practical application
- What surprised us

## Thank You!

### Questions?

---

# References (if needed)
- Data source
- R packages used
- Additional resources