---
title: "Monthly Import Statistics from Customs Data"
subtitle: "Pakistan Bureau of Statistics — Trade Statistics Wing"
author: "Prof. Dr. Zahid Asghar, Member Economic Statistics"
date: "`r format(Sys.Date(), '%B %Y')`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    css: pbs_style.css
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

# ── Package loading ────────────────────────────────────────────────────────────
library(tidyverse)
library(lubridate)
library(scales)
library(kableExtra)
library(plotly)
library(gt)
library(glue)

# ── Reproducibility seed ───────────────────────────────────────────────────────
set.seed(2025)
```

## Overview

This document demonstrates an **end-to-end automated pipeline** for processing
customs import declarations received by Pakistan Bureau of Statistics from the
Federal Board of Revenue (FBR). The pipeline:

1.  **Ingests** raw customs declaration records (Bill of Entry data)
2.  **Validates** and cleans exchange rates and unit values
3.  **Aggregates** to monthly totals by commodity and origin country
4.  **Produces** the official monthly import release table (millions USD)

> **Data source:** FBR Customs — Bill of Entry extract. NSO receives this as a
> delimited flat file or database extract on a monthly cut-off basis.

---

## Synthetic Data Generation {#sec-synth}

The function below mimics the structure of a real FBR customs extract. In
production, replace `generate_customs_data()` with a call to
`read_customs_extract()` (see @sec-production).

```{r generate-data}
#| label: generate-data

# ── Lookup tables ──────────────────────────────────────────────────────────────

hs2_chapters <- tibble(
  hs2 = c("10","15","27","29","30","38","39","40",
          "52","61","72","84","85","87","90"),
  description = c(
    "Cereals", "Animal/Vegetable Fats", "Mineral Fuels",
    "Organic Chemicals", "Pharmaceutical Products", "Misc. Chemicals",
    "Plastics", "Rubber", "Cotton", "Apparel (Knitted)",
    "Iron & Steel", "Machinery (General)", "Electrical Equipment",
    "Vehicles", "Optical/Medical Instruments"
  ),
  sector = c("Food","Food","Energy","Chemicals","Pharma","Chemicals",
             "Industrials","Industrials","Textiles","Textiles",
             "Metals","Machinery","Machinery","Transport","Tech")
)

origin_countries <- tibble(
  country_code = c("CHN","ARE","USA","SAU","IND","IDN","KWT","DEU",
                   "GBR","JPN","MYS","SGP","BEL","FRA","NLD"),
  country_name  = c("China","UAE","United States","Saudi Arabia","India",
                    "Indonesia","Kuwait","Germany","United Kingdom",
                    "Japan","Malaysia","Singapore","Belgium","France",
                    "Netherlands"),
  region = c("Asia","Middle East","Americas","Middle East","Asia",
             "Asia","Middle East","Europe","Europe","Asia",
             "Asia","Asia","Europe","Europe","Europe")
)

ports <- c("KHI","LHE","ISB","GWD","SKT")  # Karachi, Lahore, Islamabad, Gwadar, Sialkot

# ── Synthetic PKR/USD daily exchange rate ─────────────────────────────────────
generate_exchange_rates <- function(start = "2024-01-01",
                                    end   = "2024-12-31") {
  dates <- seq(as.Date(start), as.Date(end), by = "day")
  # Mild depreciation trend + noise
  n   <- length(dates)
  rate <- 280 + cumsum(rnorm(n, mean = 0.05, sd = 0.8))
  rate <- pmax(rate, 275)   # floor
  tibble(date = dates, pkr_usd = round(rate, 2))
}

# ── Core customs declaration generator ────────────────────────────────────────
generate_customs_data <- function(n_records = 18000,
                                  start_date = "2024-01-01",
                                  end_date   = "2024-12-31") {

  hs_sample      <- sample(seq_len(nrow(hs2_chapters)), n_records, replace = TRUE,
                            prob = c(3,2,8,4,5,3,4,2,2,2,5,7,7,6,3) / 63)
  country_sample <- sample(seq_len(nrow(origin_countries)), n_records, replace = TRUE,
                            prob = c(25,15,8,10,6,4,8,4,3,3,2,2,2,2,2) / 100)

  # Declaration dates — heavier in Q4
  weights_month <- c(7,6,7,7,8,8,9,8,9,9,10,12)
  month_probs   <- rep(weights_month / sum(weights_month), each = 1)

  raw_months <- sample(1:12, n_records, replace = TRUE, prob = month_probs)
  raw_days   <- sapply(raw_months, function(m) sample(1:28, 1))
  decl_dates <- as.Date(paste0("2024-", sprintf("%02d", raw_months),
                                "-", sprintf("%02d", raw_days)))

  # CIF value in USD — log-normal, chapter-specific scale
  base_values <- c(50,30,200,40,25,20,35,15,20,18,80,120,90,150,60)
  chapter_scale <- base_values[hs_sample]
  cif_usd <- round(rlnorm(n_records,
                           meanlog = log(chapter_scale),
                           sdlog   = 0.7), 2)

  # Quantity and unit
  qty_units <- c("MT","KG","LITER","UNIT","SQM","SET","PAIR","BOX")
  qty       <- round(runif(n_records, 0.5, 5000), 1)

  # Port of entry
  port_w <- c(55, 20, 8, 10, 7)
  port   <- sample(ports, n_records, replace = TRUE, prob = port_w / sum(port_w))

  # Assemble declaration table
  tibble(
    bill_no       = paste0("BoE-", sprintf("%06d", seq_len(n_records))),
    decl_date     = decl_dates,
    hs2           = hs2_chapters$hs2[hs_sample],
    country_code  = origin_countries$country_code[country_sample],
    port_of_entry = port,
    cif_usd       = cif_usd,
    quantity      = qty,
    unit          = sample(qty_units, n_records, replace = TRUE),
    # Introduce ~2 % dirty records for realism
    cif_usd       = ifelse(runif(n_records) < 0.02, NA_real_, cif_usd),
    country_code  = ifelse(runif(n_records) < 0.01, "UNK", country_code)
  )
}

# ── Generate data ──────────────────────────────────────────────────────────────
raw_customs   <- generate_customs_data(n_records = 18000)
exchange_rates <- generate_exchange_rates()

cat(glue("Records generated : {nrow(raw_customs):,}",
         "Date range        : {min(raw_customs$decl_date)} to {max(raw_customs$decl_date)}",
         "Missing CIF (raw) : {sum(is.na(raw_customs$cif_usd))}",
         .sep = "\n"))
```

---

## Validation & Cleaning {#sec-cleaning}

NSO applies a three-step validation protocol before computing official
aggregates: **(i)** structural checks, **(ii)** outlier flagging, and
**(iii)** missing-value imputation.

```{r validate}
#| label: validate

# ── Step 1: Structural validation ─────────────────────────────────────────────
validate_customs <- function(df) {
  df |>
    mutate(
      flag_missing_cif     = is.na(cif_usd),
      flag_unknown_country = country_code == "UNK",
      flag_future_date     = decl_date > Sys.Date(),
      flag_zero_qty        = quantity <= 0,
      flag_any             = flag_missing_cif | flag_unknown_country |
                             flag_future_date | flag_zero_qty
    )
}

validated <- validate_customs(raw_customs)

validation_summary <- validated |>
  summarise(
    total_records        = n(),
    missing_cif          = sum(flag_missing_cif),
    unknown_country      = sum(flag_unknown_country),
    future_date          = sum(flag_future_date),
    zero_qty             = sum(flag_zero_qty),
    total_flagged        = sum(flag_any),
    pct_flagged          = round(mean(flag_any) * 100, 2)
  )

validation_summary |>
  pivot_longer(everything(), names_to = "Check", values_to = "Count") |>
  gt() |>
  tab_header(title = "Validation Summary — Raw Customs Extract") |>
  cols_label(Check = "Validation Check", Count = "Count / %") |>
  tab_style(
    style = cell_fill(color = "#eef4ed"),
    locations = cells_body(rows = Check == "total_flagged")
  ) |>
  opt_stylize(style = 1)
```

```{r impute}
#| label: impute

# ── Step 2: Imputation — median CIF by HS2 chapter ───────────────────────────
hs2_medians <- validated |>
  filter(!flag_missing_cif) |>
  group_by(hs2) |>
  summarise(median_cif = median(cif_usd, na.rm = TRUE), .groups = "drop")

cleaned <- validated |>
  left_join(hs2_medians, by = "hs2") |>
  mutate(
    cif_usd_clean  = coalesce(cif_usd, median_cif),
    country_clean  = if_else(flag_unknown_country, "OTH", country_code),
    imputed        = flag_missing_cif
  ) |>
  # Drop residual unfixable records (should be ~0 here)
  filter(!is.na(cif_usd_clean)) |>
  select(-median_cif, -starts_with("flag_"))

# ── Step 3: Outlier detection — IQR method within HS2 ────────────────────────
cleaned <- cleaned |>
  group_by(hs2) |>
  mutate(
    q1   = quantile(cif_usd_clean, 0.25),
    q3   = quantile(cif_usd_clean, 0.75),
    iqr  = q3 - q1,
    outlier_flag = cif_usd_clean > (q3 + 3 * iqr)
  ) |>
  ungroup()

cat(glue(
  "Records after cleaning : {nrow(cleaned):,}",
  "Imputed records        : {sum(cleaned$imputed)}",
  "Outliers flagged (kept): {sum(cleaned$outlier_flag)}",
  .sep = "\n"
))
```

---

## Monthly Aggregation {#sec-aggregation}

The core NSO processing step: aggregate CIF values to **monthly totals in
millions of USD**, stratified by HS-2 chapter and country of origin.

```{r aggregate}
#| label: aggregate

# ── Join commodity and country metadata ───────────────────────────────────────
enriched <- cleaned |>
  left_join(hs2_chapters,      by = "hs2") |>
  left_join(origin_countries,  by = c("country_clean" = "country_code")) |>
  mutate(
    year_month = floor_date(decl_date, "month"),
    month_label = format(decl_date, "%b %Y")
  )

# ── Grand monthly totals (headline series) ────────────────────────────────────
monthly_totals <- enriched |>
  group_by(year_month) |>
  summarise(
    n_declarations  = n(),
    total_cif_musd  = sum(cif_usd_clean, na.rm = TRUE) / 1e6,
    .groups = "drop"
  ) |>
  mutate(
    month_label = format(year_month, "%b-%Y"),
    mom_change  = (total_cif_musd / lag(total_cif_musd) - 1) * 100,
    ytd_musd    = cumsum(total_cif_musd)
  )

# ── By HS-2 sector ────────────────────────────────────────────────────────────
monthly_by_sector <- enriched |>
  group_by(year_month, sector) |>
  summarise(cif_musd = sum(cif_usd_clean) / 1e6, .groups = "drop")

# ── By origin region ──────────────────────────────────────────────────────────
monthly_by_region <- enriched |>
  mutate(region = replace_na(region, "Other")) |>
  group_by(year_month, region) |>
  summarise(cif_musd = sum(cif_usd_clean) / 1e6, .groups = "drop")

# ── By port of entry ──────────────────────────────────────────────────────────
monthly_by_port <- enriched |>
  group_by(year_month, port_of_entry) |>
  summarise(cif_musd = sum(cif_usd_clean) / 1e6, .groups = "drop")
```

---

## Official Release Table {#sec-table}

```{r release-table}
#| label: release-table

monthly_totals |>
  select(month_label, n_declarations, total_cif_musd, mom_change, ytd_musd) |>
  gt() |>
  tab_header(
    title    = md("**Pakistan Imports — Monthly Summary**"),
    subtitle = md("*CIF Value, Millions USD · Reference Year 2024*")
  ) |>
  cols_label(
    month_label    = "Month",
    n_declarations = "Declarations",
    total_cif_musd = "CIF (M USD)",
    mom_change     = "MoM Δ (%)",
    ytd_musd       = "YTD (M USD)"
  ) |>
  fmt_number(columns = c(total_cif_musd, ytd_musd), decimals = 1) |>
  fmt_number(columns = n_declarations, decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = mom_change, decimals = 1, suffix = "%") |>
  tab_style(
    style = cell_fill(color = "#1a4a2e"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#eef4ed"),
    locations = cells_body(rows = seq(1, 12, 2))
  ) |>
  tab_style(
    style = cell_text(color = "#c0392b", weight = "bold"),
    locations = cells_body(columns = mom_change,
                           rows = mom_change < 0)
  ) |>
  tab_style(
    style = cell_text(color = "#1a4a2e", weight = "bold"),
    locations = cells_body(columns = mom_change,
                           rows = mom_change > 0)
  ) |>
  tab_source_note("Source: FBR Customs Extract processed by PBS Trade Statistics Wing") |>
  opt_stylize(style = 1)
```

---

## Charts {#sec-charts}

### Monthly Import Trend

```{r chart-trend}
#| label: chart-trend
#| fig-height: 4.5

p1 <- monthly_totals |>
  plot_ly(x = ~month_label, y = ~total_cif_musd, type = "bar",
          marker = list(color = "#1a4a2e",
                        line = list(color = "#b89a3b", width = 1)),
          name = "Monthly CIF (M USD)") |>
  add_trace(y = ~ytd_musd / 12, type = "scatter", mode = "lines+markers",
            name = "Monthly Avg (YTD)",
            line = list(color = "#b89a3b", width = 2, dash = "dash"),
            marker = list(color = "#b89a3b", size = 6)) |>
  layout(
    title  = list(text = "Pakistan Monthly Imports 2024 (CIF, M USD)",
                  font = list(family = "Georgia", size = 16, color = "#1a4a2e")),
    xaxis  = list(title = "", tickangle = -30,
                  tickfont = list(family = "Georgia")),
    yaxis  = list(title = "Millions USD",
                  titlefont = list(family = "Georgia")),
    legend = list(orientation = "h", x = 0, y = -0.25),
    plot_bgcolor  = "#f5f0e6",
    paper_bgcolor = "#f5f0e6",
    font = list(family = "Georgia")
  )

p1
```

### Imports by Sector

```{r chart-sector}
#| label: chart-sector
#| fig-height: 4.5

sector_colors <- c(
  Energy = "#1a4a2e", Machinery = "#2e7d52", Chemicals = "#b89a3b",
  Food = "#8b6914", Metals = "#4a7a62", Transport = "#c0392b",
  Pharma = "#6c3483", Tech = "#1a5276", Textiles = "#784212",
  Industrials = "#515a5a"
)

p2 <- monthly_by_sector |>
  plot_ly(x = ~format(year_month, "%b-%Y"), y = ~cif_musd, color = ~sector,
          colors = sector_colors,
          type = "bar") |>
  layout(
    barmode = "stack",
    title   = list(text = "Imports by Sector — Monthly Stack (M USD)",
                   font = list(family = "Georgia", size = 16, color = "#1a4a2e")),
    xaxis   = list(title = "", tickangle = -30),
    yaxis   = list(title = "Millions USD"),
    legend  = list(orientation = "h", x = 0, y = -0.3),
    plot_bgcolor  = "#f5f0e6",
    paper_bgcolor = "#f5f0e6",
    font = list(family = "Georgia")
  )

p2
```

### Imports by Origin Region

```{r chart-region}
#| label: chart-region
#| fig-height: 4.5

region_annual <- enriched |>
  mutate(region = replace_na(region, "Other")) |>
  group_by(region) |>
  summarise(cif_musd = sum(cif_usd_clean) / 1e6, .groups = "drop") |>
  arrange(desc(cif_musd))

p3 <- region_annual |>
  plot_ly(labels = ~region, values = ~cif_musd,
          type = "pie",
          textinfo = "label+percent",
          marker = list(colors = c("#1a4a2e","#b89a3b","#2e7d52",
                                   "#8b6914","#c0392b","#6c3483"),
                        line = list(color = "#f5f0e6", width = 1.5))) |>
  layout(
    title = list(text = "Annual Imports by Origin Region 2024",
                 font = list(family = "Georgia", size = 16, color = "#1a4a2e")),
    plot_bgcolor  = "#f5f0e6",
    paper_bgcolor = "#f5f0e6",
    font = list(family = "Georgia")
  )

p3
```

### Top 10 Partner Countries

```{r chart-countries}
#| label: chart-countries
#| fig-height: 4.5

top_countries <- enriched |>
  mutate(country_name = replace_na(country_name, "Other")) |>
  group_by(country_name) |>
  summarise(cif_musd = sum(cif_usd_clean) / 1e6, .groups = "drop") |>
  slice_max(cif_musd, n = 10) |>
  mutate(country_name = fct_reorder(country_name, cif_musd))

p4 <- top_countries |>
  plot_ly(x = ~cif_musd, y = ~country_name, type = "bar",
          orientation = "h",
          marker = list(
            color = ~cif_musd,
            colorscale = list(c(0,"#b89a3b"), c(1,"#1a4a2e")),
            line = list(color = "#f5f0e6", width = 0.5)
          )) |>
  layout(
    title = list(text = "Top 10 Import Origins — Annual 2024 (M USD)",
                 font = list(family = "Georgia", size = 16, color = "#1a4a2e")),
    xaxis = list(title = "Millions USD"),
    yaxis = list(title = ""),
    plot_bgcolor  = "#f5f0e6",
    paper_bgcolor = "#f5f0e6",
    font = list(family = "Georgia")
  )

p4
```

---

## Port-wise Breakdown {#sec-ports}

```{r port-table}
#| label: port-table

port_labels <- c(KHI = "Karachi", LHE = "Lahore", ISB = "Islamabad",
                 GWD = "Gwadar", SKT = "Sialkot")

enriched |>
  group_by(port_of_entry) |>
  summarise(
    declarations = n(),
    cif_musd     = round(sum(cif_usd_clean) / 1e6, 1),
    avg_cif_usd  = round(mean(cif_usd_clean), 0),
    share_pct    = NA_real_,
    .groups = "drop"
  ) |>
  mutate(
    share_pct  = round(cif_musd / sum(cif_musd) * 100, 1),
    port_name  = port_labels[port_of_entry]
  ) |>
  arrange(desc(cif_musd)) |>
  select(port_name, declarations, cif_musd, avg_cif_usd, share_pct) |>
  gt() |>
  tab_header(
    title    = md("**Imports by Port of Entry — 2024**"),
    subtitle = "CIF Value in Millions USD"
  ) |>
  cols_label(
    port_name    = "Port",
    declarations = "Declarations",
    cif_musd     = "CIF (M USD)",
    avg_cif_usd  = "Avg CIF (USD)",
    share_pct    = "Share (%)"
  ) |>
  fmt_number(columns = c(cif_musd), decimals = 1) |>
  fmt_number(columns = avg_cif_usd, decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = declarations, decimals = 0, use_seps = TRUE) |>
  data_color(
    columns  = share_pct,
    palette  = c("#eef4ed", "#1a4a2e")
  ) |>
  tab_source_note("Source: FBR Customs Extract · PBS Trade Statistics Wing") |>
  opt_stylize(style = 1)
```

---

## Production Integration {#sec-production}

Replace the synthetic generator with this function to connect to the real FBR
extract. The rest of the pipeline is unchanged.

```{r production-stub}
#| label: production-stub
#| eval: false

# ── Option A: CSV extract from shared drive ────────────────────────────────────
read_customs_extract <- function(path) {
  read_csv(path,
    col_types = cols(
      bill_no       = col_character(),
      decl_date     = col_date(format = "%d/%m/%Y"),
      hs2           = col_character(),
      country_code  = col_character(),
      port_of_entry = col_character(),
      cif_usd       = col_double(),
      quantity      = col_double(),
      unit          = col_character()
    )
  )
}

# ── Option B: ODBC connection to PBS/FBS Oracle database ─────────────────────
library(DBI)
library(odbc)

con <- dbConnect(odbc::odbc(),
  dsn      = "PBS_FBS_PROD",
  uid      = Sys.getenv("PBS_DB_USER"),
  pwd      = Sys.getenv("PBS_DB_PASS")
)

raw_customs <- dbGetQuery(con, "
  SELECT bill_no, decl_date, hs2_code AS hs2,
         country_code, port_code AS port_of_entry,
         cif_usd, quantity, uom AS unit
  FROM   customs.bill_of_entry
  WHERE  decl_date BETWEEN :start AND :end
", params = list(start = "2024-01-01", end = "2024-12-31"))

dbDisconnect(con)
```

---

## Headline Figures {#sec-headline}

```{r headline}
#| label: headline

annual_total <- sum(monthly_totals$total_cif_musd)
peak_month   <- monthly_totals |> slice_max(total_cif_musd, n = 1)
top_sector   <- monthly_by_sector |>
                  group_by(sector) |>
                  summarise(total = sum(cif_musd)) |>
                  slice_max(total, n = 1)
top_region   <- monthly_by_region |>
                  group_by(region) |>
                  summarise(total = sum(cif_musd)) |>
                  slice_max(total, n = 1)

tibble(
  Indicator = c("Annual Total Imports (2024)",
                "Peak Import Month",
                "Peak Month Value",
                "Largest Import Sector",
                "Largest Origin Region"),
  Value = c(
    paste0("USD ", round(annual_total, 0), " million"),
    peak_month$month_label,
    paste0("USD ", round(peak_month$total_cif_musd, 0), " million"),
    top_sector$sector,
    top_region$region
  )
) |>
  gt() |>
  tab_header(title = md("**2024 Import Headline Indicators**")) |>
  tab_style(
    style = list(cell_fill(color = "#1a4a2e"), cell_text(color = "white", weight = "bold")),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#eef4ed"),
    locations = cells_body(rows = seq(1, 5, 2))
  ) |>
  opt_stylize(style = 1)
```

---

*Document generated automatically by the PBS Trade Statistics Wing processing
pipeline. For methodology queries contact the Economic Statistics Division.*
