---
title: "Fertility Analysis: Pakistan Demographic Survey"
subtitle: "Exploring Reproductive Health Indicators"
author: "Group Names"
date: today
format:
  revealjs:
    theme: simple
    transition: slide
    slide-number: true
---

# Introduction

## Research Focus
- Analyzing fertility patterns in Pakistan
- Dataset: Pakistan Demographic Survey
- Understanding factors affecting fertility decisions

---

# Data & Setup
```{r}
#| echo: true
#| message: false

library(tidyverse)
library(readxl)  # if Excel format

# Load Pakistan Demographic Survey data
#pds_data <- read_csv("pakistan_demographic_survey.csv")
# Or: read_excel("pakistan_demographic_survey.xlsx")

# Preview data structure
glimpse(pds_data)
```

---

# Exploration Points

## 1. Total Fertility Rate by Province

**Variables to explore:**
- Province (Punjab, Sindh, KPK, Balochistan, GB, AJK)
- Total children ever born
- Age groups of women

**Analysis:**
```{r}
#| eval: false

fertility_by_province <- pds_data |>
  filter(age >= 35 & age <= 49) |>  # Completed fertility
  group_by(province) |>
  summarize(
    mean_children = mean(children_ever_born, na.rm = TRUE),
    total_women = n()
  )

# Visualize provincial differences
ggplot(fertility_by_province, aes(x = province, y = mean_children)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Average Children Born by Province",
       x = "Province", y = "Mean Number of Children")
```

---

## 2. Urban vs Rural Fertility Differentials

**Variables:**
- Residence type (urban/rural)
- Total children ever born
- Age of women

**Key Questions:**
- How does fertility differ between urban and rural areas?
- Is the gap consistent across all provinces?
```{r}
#| eval: false

pds_data |>
  filter(age >= 15 & age <= 49) |>
  group_by(residence_type, age_group) |>
  summarize(mean_children = mean(children_ever_born, na.rm = TRUE)) |>
  ggplot(aes(x = age_group, y = mean_children, 
             color = residence_type, group = residence_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Fertility by Age: Urban vs Rural")
```

---

## 3. Education and Fertility Relationship

**Variables:**
- Education level (no education, primary, secondary, higher)
- Total children ever born
- Marital status

**Analysis Focus:**
- Does education reduce fertility?
- Compare across different education levels
- Control for age and marital status
```{r}
#| eval: false

pds_data |>
  filter(marital_status == "Currently married", age >= 30) |>
  group_by(education_level) |>
  summarize(
    mean_children = mean(children_ever_born, na.rm = TRUE),
    median_children = median(children_ever_born, na.rm = TRUE),
    sample_size = n()
  ) |>
  ggplot(aes(x = education_level, y = mean_children)) +
  geom_col(fill = "darkgreen") +
  geom_text(aes(label = round(mean_children, 2)), vjust = -0.5)
```

---

## 4. Age at First Marriage and Fertility

**Variables:**
- Age at first marriage
- Total children ever born
- Current age
- Province/District

**Key Insights:**
- Does early marriage lead to higher fertility?
- Regional variations in marriage age
- Impact on women's completed fertility
```{r}
#| eval: false

pds_data |>
  filter(!is.na(age_at_marriage), age >= 40) |>
  mutate(
    marriage_category = case_when(
      age_at_marriage < 18 ~ "Before 18",
      age_at_marriage >= 18 & age_at_marriage < 25 ~ "18-24",
      age_at_marriage >= 25 ~ "25 and above"
    )
  ) |>
  group_by(marriage_category) |>
  summarize(mean_children = mean(children_ever_born, na.rm = TRUE)) |>
  ggplot(aes(x = marriage_category, y = mean_children)) +
  geom_col(fill = "coral") +
  labs(title = "Fertility by Age at Marriage",
       x = "Age at Marriage", y = "Mean Children Born")
```

---

## 5. Child Sex Preference and Fertility

**Variables:**
- Number of sons
- Number of daughters
- Total children ever born
- Desired number of children

**Analysis:**
- Do families continue having children until sons are born?
- Sex ratio at different birth orders
- Impact of son preference on fertility
```{r}
#| eval: false

# Analyze continued childbearing after sons
pds_data |>
  filter(children_ever_born >= 2) |>
  mutate(
    has_son = ifelse(number_of_sons > 0, "Has Son", "No Son")
  ) |>
  group_by(has_son) |>
  summarize(
    mean_total_children = mean(children_ever_born, na.rm = TRUE),
    continuing_fertility = mean(wants_more_children, na.rm = TRUE)
  )
```

---

## 6. Family Planning Knowledge and Use

**Variables:**
- Awareness of family planning methods
- Current use of contraception
- Number of children
- Desired family size

**Questions:**
- Knowledge vs practice gap
- Barriers to contraceptive use
- Unmet need for family planning
```{r}
#| eval: false

pds_data |>
  filter(age >= 15 & age <= 49, marital_status == "Currently married") |>
  group_by(knows_fp_method, uses_fp_method) |>
  summarize(
    count = n(),
    mean_children = mean(children_ever_born, na.rm = TRUE)
  ) |>
  mutate(percentage = count / sum(count) * 100)
```

---

# Positron Features to Demonstrate

## Data Exploration
- Variable explorer for quick data overview
- Interactive data viewer to browse rows
- Console for quick calculations

## Visualization
- Plot pane for refining graphics
- Easy export of publication-quality figures

## Code Efficiency
- Auto-completion for function names
- Integrated help and documentation
- Debugging tools

---

# Expected Findings

## Hypotheses to Test
1. Urban fertility is lower than rural fertility
2. Education inversely affects fertility
3. Early marriage increases completed fertility
4. Son preference influences fertility decisions
5. Provincial variations reflect cultural differences

---

# Conclusion

## Key Takeaways
- Summary of main pattern discovered
- Policy implications for Pakistan
- Recommendations for further analysis

## Thank You!
### Questions?