---
title: "SBP EasyData — Monthly Series Report"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: true
    fig-width: 8
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
library(tidyverse)
library(gt)
library(here)

source(here("R", "00_config.R"))

navy <- "#0f2440"
gold <- "#c9a227"

if (!file.exists(path_master)) {
  stop("No data yet -- run R/01_fetch_series.R (or run_monthly_update.R) first.")
}

df <- read_csv(path_master, show_col_types = FALSE) |>
  mutate(observation_date = as.Date(observation_date))
```

## Snapshot

```{r snapshot}
df |>
  summarise(
    latest_date = max(observation_date),
    .by = label
  ) |>
  left_join(
    df |> slice_max(observation_date, by = label) |>
      select(label, observation_value),
    by = "label"
  ) |>
  gt() |>
  cols_label(
    label              = "Series",
    latest_date        = "Latest observation",
    observation_value  = "Latest value"
  ) |>
  fmt_number(columns = observation_value, decimals = 2) |>
  tab_style(
    style = cell_fill(color = navy),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  )
```

## Trends

```{r trends, results='asis'}
for (lbl in unique(df$label)) {
  cat("###", lbl, "\n\n")

  p <- df |>
    filter(label == lbl) |>
    ggplot(aes(observation_date, observation_value)) +
    geom_line(colour = navy, linewidth = 1) +
    geom_point(colour = gold, size = 1.5) +
    labs(x = NULL, y = NULL, title = lbl) +
    theme_minimal(base_family = "Georgia") +
    theme(plot.title = element_text(face = "bold", colour = navy))

  print(p)
  cat("\n\n")
}
```

## Data notes

- Source: SBP EasyData API (`https://easydata.sbp.org.pk`), one series
  per row in `R/00_config.R`'s `series_registry`.
- Each run fetches only observations newer than what's already in
  `sbp_series_master.csv` (or backfills from `default_start_date` for a
  brand-new series) -- never duplicates.
- The API's response column names aren't documented publicly; the fetch
  script detects the date/value columns by pattern rather than assuming
  exact names, and audits what it finds on every run (see console output
  of `01_fetch_series.R`).
