---
title: "Demographic Analysis in R using DHS Model Datasets"
author: "Zahid Asghar"
date: "2025-11-19"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  freeze: auto 
editor_options: 
  chunk_output_type: console
---

# 🧭 Session 1: Introduction to DHS Data and R Setup

## Objectives
- Understand DHS data structure and variable naming
- Load and explore a DHS model dataset
- Prepare your R environment for analysis

```{r}
# Efficient package management using pacman (no pollster dependency)
if(!require(pacman)) install.packages("pacman")

pacman::p_load(
  tidyverse, haven, here, labelled, survey, broom, corrplot,
  naniar, gt, expss, xlsx
)

topline <- function(data, var, wt){
  data %>%
    dplyr::filter(!is.na({{ var }}), !is.na({{ wt }})) %>%
    dplyr::count({{ var }}, wt = {{ wt }}, name = "weight") %>%
    dplyr::mutate(weighted_pct = round(100 * weight / sum(weight), 2)) %>%
    dplyr::rename(category = {{ var }}) %>%
    dplyr::select(category, weighted_pct)
}
```

## Step 2. Load the Model Dataset

```{r}
# Load the Women's (IR) file
pkir <- read_dta(here("data", "PKIR71FL.dta"))

# Save a copy for faster loading in future sessions
save(pkir, file = "data/pkir.RData")

# Reload saved version
load("data/pkir.RData")

# Explore structure and variables
str(pkir)
```

## 🧩 Exercise 1
Inspect one variable's category labels.
```{r}
print_labels(pkir$v106)  # Education level
```

# 🧩 Session 2: Recoding Variables

```{r}
print_labels(pkir$v313)  # Check method type codes

pkir <- pkir %>%
  mutate(modfp = case_when(
    v313 == 3 ~ 1,
    v313 < 3 ~ 0
  )) %>%
  set_value_labels(modfp = c("Yes" = 1, "No" = 0)) %>%
  set_variable_labels(modfp = "Currently used any modern method")
```

```{r}
pkir <- pkir %>%
  mutate(age_child = v008 - b3_01,
         ancvisits = case_when(
           m14_1 == 0 ~ 0,
           m14_1 == 1 ~ 1,
           m14_1 %in% c(2,3) ~ 2,
           m14_1 >= 4 & m14_1 <= 90 ~ 3,
           TRUE ~ NA_real_
         )) %>%
  set_value_labels(ancvisits = c("None" = 0, "1" = 1, "2-3" = 2, "4+" = 3)) %>%
  set_variable_labels(ancvisits = "Number of ANC visits")
```

# ⚖️ Session 3: Understanding and Applying Survey Weights

## Why Use Weights?
DHS employs complex sampling. Weights adjust for unequal selection probabilities and ensure representativeness.

```{r}
pkir <- pkir %>%
  mutate(wt = v005 / 1e6)
```

```{r}
topline(pkir, modfp, wt)
```

# 🧮 Session 4: Defining the Survey Design Object

```{r}
mysurvey <- svydesign(
  id = pkir$v021,
  strata = pkir$v022,
  weights = pkir$wt,
  data = pkir,
  nest = TRUE
)
options(survey.lonely.psu = "adjust")
```

```{r}
prop.table(svytable(~modfp, mysurvey))
```

# 📊 Session 5: Crosstabulations and Associations

```{r}
table_edu <- svyby(~modfp, by = ~v106, design = mysurvey, FUN = svymean, vartype = c("se", "ci"))
table_edu

svychisq(~modfp + v106, mysurvey)
```

# 📈 Session 6: Logistic Regression Analysis

```{r}
reg1 <- svyglm(modfp ~ v025, design = mysurvey, family = binomial(link = "logit"))
summary(reg1)

exp(reg1$coefficients)
```

```{r}
reg2 <- svyglm(modfp ~ v025 + as.factor(v106) + as.factor(v013) + as.factor(v190),
               design = mysurvey, family = binomial(link = "logit"))

library(broom)
tidy(reg2, exponentiate = TRUE, conf.int = TRUE)
```

# 🧮 Session 7: Correlation and Diagnostics

```{r}
cordata <- pkir %>% select(v025, v106, v013, v190)
cor_matrix <- cor(cordata, use = "pairwise.complete.obs")

corrplot(cor_matrix, method = "circle")
```

# 📤 Session 8: Exporting and Presenting Results

```{r, eval=FALSE}
write.xlsx(table_edu, "outputs/weighted_tables.xlsx", sheetName = "Education vs FP")
```

```{r}
top_tbl <- topline(pkir, modfp, wt) %>%
  mutate(category = as.character(category)) 

# Convert from labelled to text



gt_tbl <- gt::gt(top_tbl)

gt_tbl <- gt::fmt_number(
  data = gt_tbl,
  columns = dplyr::vars(weighted_pct),
  decimals = 2
)

gt_tbl <- gt::tab_header(
  gt_tbl,
  title = gt::md("**Modern Contraceptive Use among Women (Weighted)**")
)

gt_tbl
```

# 🎓 Wrap-Up and Reflection

| Step | Concept | Function(s) |
|------|----------|-------------|
| Data loading | Importing DHS data | `read_dta()` |
| Recoding | Creating new indicators | `mutate()`, `case_when()` |
| Weighting | Adjust for complex survey | `v005 / 1e6` |
| Design setup | Declare design | `svydesign()` |
| Analysis | Weighted stats | `svymean()`, `svyby()`, `svyglm()` |
| Export | Save outputs | `write.xlsx()`, `gt()` |
