---
title: "Introduction to R"
subtitle: "Pakistan District MPI, 2019-20"
author: "D4D | Data for Development"
format:
  revealjs:
    theme: [default, d4d_theme.scss]
    slide-number: true
    transition: fade
    code-line-numbers: true
    highlight-style: github
---

## Today's Data

Pakistan District-level MPI, 2019-20 — 127 districts

Three core variables:

- **mpi** — overall poverty score (higher = poorer)
- **h_incidence** — H: % of population that is poor
- **a_intensity** — A: average deprivation among the poor

Relationship: **MPI = H × A / 100**

---

## Step 1: Load Tools

```{r}
#| eval: false
#| echo: true
library(tidyverse)
```

One package. Everything we need today lives inside it.

---

## Step 2: Import Data

```{r}
#| eval: false
#| echo: true
mpi <-  here::here("data", "pakistan_mpi_district_2019_20.csv")  # check file path)
```

`read_csv()` reads the file and stores it as `mpi`.

---

## Step 3: First Look at the Data

```{r}
#| eval: false
#| echo: true
dim(mpi)        # rows x columns
str(mpi)        # column names and types
head(mpi)       # first 6 rows
glimpse(mpi)    # compact structure view
```

Run these four commands on every new dataset.

---

## Step 4: Further Inspection

```{r}
#| eval: false
#| echo: true
tail(mpi)       # last 6 rows
summary(mpi)    # min / mean / max for numeric columns
```

---

## Step 5: Standardise Column Names

```{r}
#| eval: false
#| echo: true
mpi <- mpi |> janitor::clean_names()
names(mpi)
```

Good practice for every imported dataset, regardless of source.

---

## Step 6: Focus on the Three Core Variables

```{r}
#| eval: false
#| echo: true
mpi |>
  select(district, mpi, h_incidence, a_intensity) |>
  head(10)
```

`select()` narrows the data to the columns we need.

---

## Step 7: Rank Districts by MPI

```{r}
#| eval: false
#| echo: true
mpi |>
  select(district, province, mpi) |>
  arrange(desc(mpi)) |>
  head(10)
```

`arrange(desc())` sorts from highest to lowest.

---

## Step 8: Filter by Province

```{r}
#| eval: false
#| echo: true
mpi |>
  filter(province == "Balochistan") |>
  select(district, mpi, h_incidence, a_intensity) |>
  arrange(desc(mpi))
```

`filter()` keeps only rows meeting a condition.

---

## Step 9: Province-Level Averages

```{r}
#| eval: false
#| echo: true
mpi |>
  summarise(
    avg_mpi = mean(mpi),
    avg_h   = mean(h_incidence),
    avg_a   = mean(a_intensity),
    .by = province
  ) |>
  arrange(desc(avg_mpi))
```

`.by =` groups without a separate `group_by()` / `ungroup()` pair.

---

## Step 10: Build a Plot — The Canvas

```{r}
#| eval: false
#| echo: true
ggplot(mpi, aes(x = h_incidence, y = a_intensity))
```

No geometry mapped yet — this produces an empty canvas.

---

## Step 10: Build a Plot — Add Points

```{r}
#| eval: false
#| echo: true
ggplot(mpi, aes(x = h_incidence, y = a_intensity)) +
  geom_point()
```

Each point now represents one district.

---

## Step 10: Build a Plot — Style and Label

```{r}
#| eval: false
#| echo: true
ggplot(mpi, aes(x = h_incidence, y = a_intensity, color = province)) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title    = "District-Level MPI Components, 2019-20",
    subtitle = "Each point represents one district",
    x        = "H: Incidence of poverty (%)",
    y        = "A: Intensity of deprivation",
    color    = "Province"
  )
```

---

## Step 11: Bar Chart of the 10 Poorest Districts

```{r}
#| eval: false
#| echo: true
top10 <- mpi |>
  arrange(desc(mpi)) |>
  head(10)

ggplot(top10, aes(x = reorder(district, mpi), y = mpi, fill = province)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "10 Poorest Districts by MPI Score (2019-20)",
    x = NULL,
    y = "MPI Score"
  )
```

---

## Step 12: Export Outputs

```{r}
#| eval: false
#| echo: true
province_summary <- mpi |>
  summarise(
    avg_mpi = mean(mpi),
    avg_h   = mean(h_incidence),
    avg_a   = mean(a_intensity),
    .by = province
  )

write_csv(province_summary, "province_mpi_summary.csv")
ggsave("top10_poorest_districts.png", width = 8, height = 5)
```

---

## Summary of Commands Covered

| Task       | Commands |
|------------|----------|
| Import     | `read_csv()` |
| Inspect    | `dim()`, `str()`, `head()`, `tail()`, `glimpse()`, `summary()` |
| Clean      | `clean_names()` |
| Transform  | `select()`, `filter()`, `arrange()`, `summarise()`, `.by` |
| Visualize  | `ggplot()`, `geom_point()`, `geom_col()`, `theme_minimal()` |
| Export     | `write_csv()`, `ggsave()` |

---

## Next

Same workflow, any column in the dataset.

**Try it yourself:** repeat Steps 6–9 using `pop_share` or `n_poor_thousands`.
