---
title: "Communicating Better With Data: Import, Export & APIs in R"
subtitle: "Pakistan Social Datasets — PSLM, ASER, LFS, PDHS, MICS"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

# 1. Introduction

Data communication begins long before charts or dashboards.
It starts from one fundamental skill: **getting data in and out of your analytical environment** — cleanly, reproducibly, and transparently.

In Pakistan’s context, major datasets appear in **various formats**:

| Dataset  | Institution                   | File Format         | Notes                                  |
| -------- | ----------------------------- | ------------------- | -------------------------------------- |
| **PSLM** | Pakistan Bureau of Statistics | Excel, CSV          | Often merged from district-level files |
| **ASER** | ASER Pakistan                 | Excel, CSV          | Education indicators                   |
| **LFS**  | PBS                           | Stata (`.dta`), CSV | Microdata often obtained via request   |
| **MICS** | UNICEF                        | Stata (`.dta`)      | Standardized sampling                  |
| **PDHS** | NIPS & DHS                    | Stata (`.dta`)      | Includes household & individual files  |

This chapter teaches you how to work with these formats efficiently in **R**, with **Python** and **Stata** guidance provided in sidebars.

---

# 2. Getting Started in R

```{r}
# Load core packages
library(readr)      # CSV
library(readxl)     # Excel
library(haven)      # Stata, SPSS, SAS
library(jsonlite)   # JSON
library(httr)       # API calls
library(dplyr)      
library(tidyr)
library(ggplot2)
```

If you prefer a **one-shot loader**, you can use:

```{r}
install.packages("pacman")
pacman::p_load(readr, readxl, haven, jsonlite, httr, dplyr, tidyr, ggplot2)
```

---

# 3. Importing Data

## 3.1 Importing CSV: Example PSLM & ASER

Assume PSLM district-level file saved as:

```
data/pslm_2020_district.csv
```

```{r}
pslm <- read_csv("data/pslm_2020_district.csv")
head(pslm)
```

### ASER Example

```{r}
aser <- read_csv("data/ASER_2023_school_data.csv")
glimpse(aser)
```

---

### 📌 *Python Equivalent (sidebar)*

```python
import pandas as pd
pslm = pd.read_csv("data/pslm_2020_district.csv")
```

### 📌 *Stata Equivalent (sidebar)*

```stata
import delimited "data/pslm_2020_district.csv", clear
```

---

## 3.2 Importing Excel: PSLM & ASER

Many Pakistan datasets arrive as Excel with multiple sheets.

```{r}
pslm_xlsx <- read_excel("data/PSLM_2020.xlsx", sheet = "Districts")
head(pslm_xlsx)
```

To list the sheets:

```{r}
excel_sheets("data/PSLM_2020.xlsx")
```

---

### Example: Import ASER Village-Level File

```{r}
aser_village <- read_excel("data/ASER_2023.xlsx", sheet = 3)
glimpse(aser_village)
```

---

## 3.3 Importing Stata Data: PDHS, LFS, MICS

PBS LFS, MICS, and PDHS usually come as `.dta`.

```{r}
lfs <- haven::read_dta("data/LFS_2020.dta")
glimpse(lfs)
```

### Example: PDHS Household File

```{r}
pdhs <- read_dta("data/PKIR71FL.DTA")
summary(pdhs$v012)   # Age of respondent
```

---

### 📌 Python Sidebar

```python
import pyreadstat
lfs, meta = pyreadstat.read_dta("data/LFS_2020.dta")
```

### 📌 Stata Sidebar

```stata
use "data/LFS_2020.dta", clear
```

---

## 3.4 Importing JSON: Population Census API, Custom Sources

Assume a JSON file containing district indicators:

```json
{
  "district": "Lahore",
  "literacy_rate": 74.2,
  "sample": 12000
}
```

Load in R:

```{r}
json_data <- jsonlite::fromJSON("data/district_literacy.json")
json_data
```

---

## 3.5 APIs & Web Data Access: Example with Pakistan Open Data API

Let’s use **Open Data Pakistan** (a mock API for demonstration).

```{r}
url <- "https://api.pakopendata.pk/v1/education/pslm?district=Lahore"

response <- httr::GET(url)
content <- httr::content(response, as = "parsed")
content$data
```

Create a data frame:

```{r}
df_api <- as.data.frame(content$data)
df_api
```

---

### Real Example: Fetching JSON from a URL

```{r}
url_json <- "https://raw.githubusercontent.com/datasets/population/master/data/population.json"
pop <- jsonlite::fromJSON(url_json)
head(pop)
```

---

# 4. Exporting Data

## 4.1 Export CSV (PSLM example)

```{r}
write_csv(pslm, "output/pslm_cleaned.csv")
```

---

## 4.2 Export Excel

```{r}
library(writexl)
write_xlsx(list("pslm" = pslm, "aser" = aser), "output/pakistan_data.xlsx")
```

---

## 4.3 Export Stata

```{r}
write_dta(pslm, "output/pslm_cleaned.dta")
```

---

## 4.4 Export JSON

```{r}
jsonlite::write_json(pslm, "output/pslm_cleaned.json", pretty = TRUE)
```

---

# 5. Data Transformation & Cleaning

*(With examples from PSLM, ASER, PDHS)*

## 5.1 Cleaning District Names (PSLM)

```{r}
pslm <- pslm %>%
  mutate(
    district = stringr::str_to_title(district),
    province = stringr::str_to_title(province)
  )
```

---

## 5.2 Creating Indicators (PDHS Example)

Modern contraceptive use:

```{r}
pdhs <- pdhs %>%
  mutate(
    modern_use = ifelse(v313 == 1, 1, 0)
  )
```

---

## 5.3 Summarising Literacy by District (ASER)

```{r}
aser_summary <- aser %>%
  group_by(District) %>%
  summarise(
    literacy_rate = mean(reading_level >= 2, na.rm = TRUE)
  )
```

---

# 6. Visualizing Data for Communication

## 6.1 PSLM: Female Literacy by Province

```{r}
ggplot(pslm, aes(x = province, y = female_literacy)) +
  geom_col(fill = "#2E86C1") +
  theme_minimal() +
  labs(
    title = "Female Literacy Across Provinces — PSLM 2020",
    x = "",
    y = "%"
  )
```

---

## 6.2 ASER: Children Who Can Read a Story

```{r}
ggplot(aser_summary, aes(x = reorder(District, literacy_rate),
                         y = literacy_rate)) +
  geom_col(fill = "#D35400") +
  coord_flip() +
  labs(
    title = "ASER 2023 — Literacy (Story Level)",
    x = "District",
    y = "Proportion"
  ) +
  theme_bw()
```

---

# 7. Working With Large Datasets

### (LFS, PDHS, PSLM microdata)

Large microdata from PBS or DHS often exceed memory.

## 7.1 Use `data.table::fread`

```{r}
library(data.table)
lfs_fast <- fread("data/LFS_2020_large.csv")
```

## 7.2 Use Arrow for parquet datasets

```{r}
library(arrow)
lfs_arrow <- read_parquet("data/LFS2020.parquet")
```

---

# 8. Building Reproducible Workflows in Positron

### Key Principles:

1. **Use `here::here()` for paths**
2. **Use Quarto projects**
3. **Separate raw and cleaned data**
4. **Document everything**
5. **Use version control (Git/GitHub)**

### Example folder structure:

```
project/
 ├── data_raw/
 ├── data_clean/
 ├── scripts/
 ├── analysis/
 ├── quarto/
 └── outputs/
```

---

# 9. Summary Checklist

This chapter covered:

✔ Importing: CSV, XLSX, DTA, JSON
✔ Exporting to: CSV, XLSX, JSON, DTA
✔ APIs & Web Data
✔ Pakistan datasets (PSLM, ASER, LFS, PDHS, MICS)
✔ Cleaning & transforming data
✔ Visual communication with ggplot2
✔ Reproducible workflows in Positron

---

# 10. Exercises (Pakistan-Data Focus)

### **Exercise 1: Import & Clean PSLM**

* Import PSLM 2020 CSV file
* Clean district and province names
* Create literacy summary table

### **Exercise 2: Import PDHS Individual File**

* Load `.dta` file
* Create indicator for modern contraceptive use
* Create summary by province

### **Exercise 3: Fetch JSON from a URL**

* Use JSON API
* Convert to data frame
* Plot a bar chart

---

If you'd like, I can also provide:

✅ A **PDF Quarto version**
✅ A **full book chapter + exercises**
✅ Additional chapters (Exploratory Data Analysis, Mapping Pakistan, Dashboarding with Quarto)
✓ Or convert this into **Slides (Revealjs)**

Would you like the next chapter: **"Exploratory Data Analysis with Pakistani Social Datasets"?**
