---
title: "Working Smarter with dplyr 1.2.0"
subtitle: "Pakistan flood exposure, 2000–2024 (EM-DAT × MODIS)"
author: "Prof. Dr. Zahid Asghar"
format:
  revealjs:
    slide-number: true
    theme: simple
    code-line-numbers: false
    footer: "d4d · Data-for-Development R Training"
execute:
  echo: true
  eval: false   # flip to true once you run this locally — see note on slide 3
---

## Today's data

Disaggregated EM-DAT flood disaster records, matched to MODIS-derived flood
maps at the admin1-month level (2000–2024).

Source: Nicole Keeney, [`emdat-modis-flood-dataset`](https://github.com/nicolejkeeney/emdat-modis-flood-dataset)

```{r}
library(tidyverse)   # dplyr >= 1.2.0 (CRAN, released 2026-02-03)

flood_monthly <- readr::read_csv(
  "https://raw.githubusercontent.com/nicolejkeeney/emdat-modis-flood-dataset/refs/heads/main/data/emdat_modis_flood_dataset.csv"
)
```

`flood_monthly` has one row per **admin1-region × flood-event**, with
`flooded_population`, `flooded_area`, and normalized versions of each.

::: aside
`dplyr` 1.2.0 is needed for `filter_out()`, `when_any()`/`when_all()`, and
`recode_values()`/`replace_values()`/`replace_when()` used later.
`install.packages("dplyr")` on a machine with CRAN access, or `pak::pak("dplyr")`.
:::

## A note before we start

I don't have an R interpreter in this sandbox, so every chunk below is
**hand-verified against the actual CSVs**, not executed. Please run the deck
top-to-bottom once (`quarto render` or step through in Positron) before you
teach from it, the usual practice for any new material.

I pulled the three data files and inspected them directly (`curl` + `python3`)
to check column names, date formats, and the Pakistan subset before writing
any of the code that follows — that's where the corrections below come from.

## Is this dataset OK for Pakistan? {.smaller}

Broadly **yes for teaching disaggregation, joins, and data-quality workflows**;
use with real caveats if anyone wants to cite numbers.

- **It's disaster-scale, not comprehensive.** EM-DAT only logs events that
  cross its threshold (deaths, people affected, a declared emergency, or an
  international appeal) — routine local flooding won't be here. Treat it as
  "notable flood events," not a flood monitoring feed.
- **MODIS + monsoon clouds.** MODIS optical imagery struggles with cloud cover,
  and Pakistan's floods are dominated by the monsoon (Jul–Sep), exactly when
  cloud cover is worst. Flag 12 ("no flooded pixels detected") is common for
  Pakistan and may mean *cloud-obscured*, not *no flood* — worth teaching as a
  detection-limitation, not ground truth.
- **Mixed admin1 vintage** — see next slide, this is the big one.
- **Population**: from NASA's GPW gridded population, not Pakistan's own
  census — fine for relative exposure, not for official reporting.

## The adm1_code problem {.smaller}

Pakistan's rows use **two different administrative resolutions**:

| adm1_code range | n rows | What it is |
|---|---|---|
| 2272–2277 (6 codes) | 269 | The six GAUL "core" units: 4 provinces + ICT + FATA |
| 40408, 40409 | 54 | Two larger units — population size suggests **Azad Jammu & Kashmir** and **Gilgit-Baltistan** territory totals, but this is my inference from population magnitude, not a confirmed source |
| 40422–40431 (10 codes) | 90 | Ten smaller units, consistent with **AJK's 10 districts** by count — again inferred, not confirmed |

I verified the 2272–2277 mapping against an FAO GIEWS export that carries GAUL
codes alongside province names. I could **not** find an authoritative source
for the 40000-series codes in the time available — before publishing anything
with those regions named, pull the GAUL 2015 admin1 shapefile's attribute
table and check by hand.

One more wrinkle: **FATA (2273) merged into Khyber Pakhtunkhwa in 2018** — this
dataset predates that, so treat 2273 and 2275 as one region for anything
post-2018.

## Load the rest of the data — with two fixes

```{r}
flood_adm1_summary <- read_csv(
  "https://raw.githubusercontent.com/nicolejkeeney/emdat-modis-flood-dataset/refs/heads/main/data/adm1_summary_stats.csv"
)

# Original code pointed at a GitHub *blob* page (renders HTML, not a CSV) and
# re-read adm1_summary_stats.csv a second time under a different name.
# The file we actually want is data_processing_flags.csv:
flood_processing_flags <- read_csv(
  "https://raw.githubusercontent.com/nicolejkeeney/emdat-modis-flood-dataset/refs/heads/main/data/data_processing_flags.csv"
)

flood_processing_flags
```

```
  flag flag_meaning
  <dbl> <chr>
1     1 Start day originally NaN and was filled with first day of the month
2     2 End day originally NaN and was filled with last day of the month
...
12    No flooded pixels detected by MODIS water detection algorithm
```

## Quick review of dplyr verbs

```{r}
pakistan_floods <- flood_monthly |>
  filter(ISO == "PAK")

glimpse(pakistan_floods)

pakistan_floods |> count(ISO)
pakistan_floods |> count(adm1_code, sort = TRUE)
```

`filter()` for **keeping** rows — 415 of them, for Pakistan. Six of the eight
`adm1_code` groupings above account for most of that count.

## The `mon-yr` bug {.smaller}

```{r}
pakistan_floods |> distinct(`mon-yr`) |> slice_sample(n = 6)
```

```
# A tibble: 6 × 1
  `mon-yr`
  <chr>
1 1-Jul
2 2-Aug
3 10-Aug
4 22-Aug
5 3-Jul
6 5-Sep
```

That's **not** `Jan-00` (month-year) — it's `D-Mon`. Somewhere upstream, a
spreadsheet auto-converted `"Jul-01"` into a date and re-exported it, silently
losing the year for 97.6% of all rows in the full dataset (22,794 of 23,344).
`lubridate::ymd()` on this returns `NA` with a parse-failure warning for
almost every Pakistan row — a silent, easy-to-miss bug, not a loud one.

**Fix: don't use `mon-yr` at all.** `start_date`/`end_date` are consistently
`M/D/YY` across all 23,344 rows — build `month` from `start_date` instead.

## Corrected date handling

```{r}
pakistan_floods <- pakistan_floods |>
  mutate(
    start_date = lubridate::mdy(start_date),
    end_date   = lubridate::mdy(end_date),
    month      = lubridate::floor_date(start_date, unit = "month"),
    .keep = "unused"          # drop the unreliable `mon-yr` column
  )

pakistan_floods |>
  select(start_date, end_date, month) |>
  head()
```

Note the **order**: `start_date` has to be parsed *before* `month` can be
derived from it — the original code computed `month` from `mon-yr` in the same
`mutate()` call where `start_date` was still character text.

## Nice names, the dplyr 1.2.0 way

`recode_values()` replaces the old `recode()` / verbose `case_when()` chain
for mapping codes onto labels:

```{r}
pakistan_provinces <- tibble::tribble(
  ~adm1_code, ~province_name,
  2272,  "Balochistan",
  2273,  "FATA (merged into KP, 2018)",
  2274,  "Islamabad Capital Territory",
  2275,  "Khyber Pakhtunkhwa",
  2276,  "Punjab",
  2277,  "Sindh",
  40408, "Azad Jammu & Kashmir (territory total)*",
  40409, "Gilgit-Baltistan (territory total)*",
  # 40422–40431: AJK districts — codes not yet confirmed, see slide 5
)

pakistan_floods <- pakistan_floods |>
  mutate(adm1_code = as.numeric(adm1_code)) |>
  mutate(
    province_name = recode_values(
      adm1_code,
      from = pakistan_provinces$adm1_code,
      to   = pakistan_provinces$province_name,
      default = as.character(adm1_code),
      unmatched = "default"  # keeps the raw code for the unresolved districts
    )
  )

pakistan_floods |> count(province_name, sort = TRUE)
```

## Dropping rows: `filter()` vs `filter_out()`

The `flags` column is semicolon-separated and multi-valued (e.g. `"1; 2; 7"`).
Flag **6** — *"coordinate mismatch error between floodmap and GPW population
data"* — genuinely signals an unreliable row, unlike flag 12 (which, per
slide 4, plausibly reflects real MODIS/cloud limitations rather than bad data).

```{r}
# Old way — keeping requires the negative case, and you have to remember
# str_detect() so you don't accidentally drop everything with an NA flags cell:
pakistan_floods |>
  filter(!str_detect(flags, "\\b6\\b") | is.na(flags))

# dplyr 1.2.0 way — filter_out() is built for dropping, NAs are kept by default:
pakistan_floods_clean <- pakistan_floods |>
  filter_out(str_detect(flags, "\\b6\\b"))
```

## Combining conditions: `when_any()` / `when_all()`

"High-exposure" flood-months: **Sindh or Punjab** with >1% of population
flooded, **or** any region with a genuinely large flooded population.

```{r}
high_exposure <- pakistan_floods_clean |>
  filter(
    when_any(
      province_name %in% c("Sindh", "Punjab") & flooded_population_norm > 0.01,
      flooded_population > 50000
    )
  )

high_exposure |> count(province_name, sort = TRUE)
```

Reads the same way you'd say it out loud — no `|`/`&` precedence puzzles, and
no accidental NA-dropping the way plain `filter()` can produce.

## Monthly aggregation — fixing a double-count

```{r}
# One region can have >1 distinct EM-DAT event in the same month (9 cases for
# Pakistan). total_population is a static per-region figure, repeated on every
# row — summing it across overlapping events inflates it. flooded_population
# is genuinely event-specific and safe to sum.

pakistan_region_month <- pakistan_floods_clean |>
  summarise(
    flooded_population = sum(flooded_population, na.rm = TRUE),
    flooded_area        = sum(flooded_area, na.rm = TRUE),
    total_population     = first(total_population),   # not summed
    .by = c(adm1_code, province_name, month)
  )

pakistan_monthly <- pakistan_region_month |>
  summarise(
    flooded_population = sum(flooded_population, na.rm = TRUE),
    flooded_area        = sum(flooded_area, na.rm = TRUE),
    total_population     = sum(total_population, na.rm = TRUE),  # safe now
    .by = month
  )
```

## Plot 1 — monthly exposure by region

```{r}
ggplot(pakistan_region_month, aes(month, flooded_population)) +
  geom_line() +
  facet_wrap(~ province_name, scales = "free_y") +
  labs(
    title = "Monthly flood exposure by administrative region",
    subtitle = "Pakistan, EM-DAT × MODIS, 2000–2024",
    x = NULL,
    y = "Flooded population (MODIS-derived)",
    caption = "Source: emdat-modis-flood-dataset (Keeney). *Territory-level codes inferred, see notes."
  )
```

`province_name` reads far better in a facet strip than a bare `adm1_code`.

## Plot 2 — normalized exposure

```{r}
ggplot(pakistan_floods_clean, aes(month, flooded_population_norm)) +
  geom_line() +
  facet_wrap(~ province_name) +
  labs(
    title = "Normalized monthly flood exposure",
    subtitle = "Share of regional population flooded",
    x = NULL,
    y = "Flooded population ÷ total population"
  )
```

`flooded_population_norm` already lives in `pakistan_floods` — no join to
`flood_adm1_summary` needed for this one. That summary table is still useful
for a *per-region average across all 24 years* (`event_count`,
`mean_flooded_population`, …), just not for this monthly view.

## What changed, in one place {.smaller}

1. `flood_processing_flags` — wrong URL (HTML blob page) *and* wrong file;
   now reads `data_processing_flags.csv` via `raw.githubusercontent.com`.
2. `month = ymd(\`mon-yr\`)` silently returned `NA` for ~98% of rows because
   `mon-yr` is corrupted (`"Jul-01"` → `"1-Jul"`, year lost). Now derived from
   `start_date` after parsing it.
3. Parse order fixed: dates before deriving `month` from them.
4. Added `province_name` via `recode_values()` — nice names, dplyr 1.2.0 style.
5. Added `filter_out()` to drop coordinate-mismatch rows (flag 6) explicitly,
   keeping `NA`-flag rows rather than silently dropping them.
6. Fixed a real double-count: `total_population` was being summed across
   overlapping events in the same region-month.
7. Switched `group_by() |> summarise(.groups = "drop")` to `.by =`.

## References

- Keeney, N. *EM-DAT MODIS Flood Dataset*. github.com/nicolejkeeney/emdat-modis-flood-dataset
- Vaughan, D. & Posit team. *dplyr 1.2.0*. opensource.posit.co, 2026-02-04
- Velásquez, I. *Working Smarter with dplyr 1.2.0*, R-Ladies Rome
- `data_processing_flags.csv` for the full flag legend (12 codes)
