---
title: "SDPI D4D Workshop — Pre-Training Survey"
subtitle: "Data Analysis Using R for Evidence-Based Decision-Making, 31 Aug – 4 Sep 2026"
author: "Data for Development (D4D), School of Economics, QAU"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    code-fold: true
    embed-resources: true
    fig-width: 8
    fig-height: 4.5
params:
  sheet_id: "1Ttg8BfByE9BEW7fy0YtLtWwBY7VmV7V0COdr0EgUjNc"
  gid: "1358570388"
  use_auth: false
execute:
  warning: false
  message: false
  echo: false
---

<!--
SETUP (one time)

Option A — no authentication (recommended for a live report)
  In the responses Google Sheet: File > Share > Share with others
  > General access > "Anyone with the link" > Viewer.
  Leave params$use_auth = false. Re-render whenever you want fresh numbers.

Option B — private sheet
  install.packages("googlesheets4"); set params$use_auth: true
  First render will open a browser for Google sign-in and cache the token.

The report reads whatever columns exist in the sheet and classifies each one
as single-choice, multi-select (checkbox), or free text. New form questions
appear automatically on the next render; no edits needed here.
-->

```{r}
#| label: setup
library(tidyverse)
library(janitor)
library(knitr)

navy <- "#0f2440"
gold <- "#c9a227"
```

```{r}
#| label: read-data

sheet_url <- str_glue(
  "https://docs.google.com/spreadsheets/d/{params$sheet_id}",
  "/export?format=csv&gid={params$gid}"
)

raw <- if (isTRUE(params$use_auth)) {
  googlesheets4::read_sheet(params$sheet_id, sheet = NULL, col_types = "c")
} else {
  read_csv(sheet_url, show_col_types = FALSE, col_types = cols(.default = col_character()))
}

# Keep original question wording for headings; use clean names for handling.
question_labels <- names(raw)
responses <- raw |>
  clean_names() |>
  filter(if_any(everything(), \(x) !is.na(x) & str_trim(x) != ""))

names(question_labels) <- names(responses)
```

```{r}
#| label: classify-columns

# Timestamp -------------------------------------------------------------
ts_col <- names(responses)[str_detect(names(responses), "timestamp|time_stamp|submitted")] |>
  head(1)

if (length(ts_col) == 1) {
  responses <- responses |>
    mutate(
      .submitted = parse_date_time(
        .data[[ts_col]],
        orders = c("mdY HMS", "mdY HM", "dmY HMS", "Ymd HMS", "mdY"),
        quiet = TRUE
      )
    )
}

# Drop identifiers that should not be charted --------------------------
drop_pattern <- "timestamp|time_stamp|submitted|email|e_mail|score|^name$|full_name|phone|cnic|contact"
candidate_cols <- setdiff(
  names(responses)[!str_detect(names(responses), drop_pattern)],
  ".submitted"
)

# Classify each remaining column ---------------------------------------
classify_column <- function(x) {
  x <- str_trim(as.character(x))
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return("empty")

  n_unique   <- n_distinct(x)
  mean_len   <- mean(str_length(x))
  comma_frac <- mean(str_detect(x, ","))
  split_vals <- str_split(x, ",\\s*") |> unlist() |> str_trim() |> unique()

  if (mean_len > 60 || (n_unique >= 0.9 * length(x) && mean_len > 30)) return("text")
  if (comma_frac > 0.15 && length(split_vals) < n_unique) return("multi")
  "single"
}

col_type <- map_chr(candidate_cols, \(v) classify_column(responses[[v]]))
names(col_type) <- candidate_cols

single_cols <- names(col_type)[col_type == "single"]
multi_cols  <- names(col_type)[col_type == "multi"]
text_cols   <- names(col_type)[col_type == "text"]
```

```{r}
#| label: helpers

# Common Likert / capability scales, ordered low -> high.
known_scales <- list(
  c("None", "Beginner", "Basic", "Intermediate", "Advanced", "Expert"),
  c("Never", "Rarely", "Sometimes", "Often", "Always"),
  c("Strongly disagree", "Disagree", "Neutral", "Agree", "Strongly agree"),
  c("Not at all confident", "Slightly confident", "Moderately confident",
    "Very confident", "Extremely confident"),
  c("Not important", "Slightly important", "Moderately important",
    "Very important", "Extremely important"),
  c("Not at all", "A little", "Somewhat", "Quite a lot", "A great deal"),
  c("Low", "Medium", "High"),
  c("Yes", "No", "Not sure", "Maybe")
)

order_levels <- function(values, counts) {
  vals_lower <- str_to_lower(values)
  for (scale in known_scales) {
    if (all(vals_lower %in% str_to_lower(scale))) {
      return(scale[str_to_lower(scale) %in% vals_lower])
    }
  }
  values[order(counts, decreasing = FALSE)]  # frequency order for bar chart
}

tally_single <- function(data, col) {
  data |>
    transmute(answer = str_trim(.data[[col]])) |>
    filter(!is.na(answer), answer != "") |>
    count(answer, name = "n") |>
    mutate(
      pct    = n / sum(n),
      answer = factor(answer, levels = order_levels(answer, n))
    ) |>
    arrange(desc(n))
}

tally_multi <- function(data, col) {
  n_resp <- data |>
    filter(!is.na(.data[[col]]), str_trim(.data[[col]]) != "") |>
    nrow()

  data |>
    transmute(answer = str_split(.data[[col]], ",\\s*")) |>
    unnest(answer) |>
    mutate(answer = str_trim(answer)) |>
    filter(!is.na(answer), answer != "") |>
    count(answer, name = "n") |>
    mutate(
      pct    = n / n_resp,
      answer = factor(answer, levels = answer[order(n)])
    ) |>
    arrange(desc(n))
}

plot_tally <- function(tbl, subtitle = NULL) {
  ggplot(tbl, aes(x = n, y = answer)) +
    geom_col(fill = navy, width = 0.65) +
    geom_text(
      aes(label = str_glue("{n}  ({scales::percent(pct, accuracy = 1)})")),
      hjust = -0.15, size = 3.4, colour = navy
    ) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.22))) +
    scale_y_discrete(labels = \(x) str_wrap(x, 45)) +
    labs(x = "Respondents", y = NULL, subtitle = subtitle) +
    theme_minimal(base_size = 12, base_family = "Georgia") +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      axis.text.y        = element_text(colour = navy),
      plot.subtitle      = element_text(colour = gold, face = "bold")
    )
}
```

## Overview

```{r}
#| label: overview

n_responses <- nrow(responses)

overview <- tibble(
  Item = c("Responses received", "Questions in form", "Report generated"),
  Value = c(
    as.character(n_responses),
    as.character(length(candidate_cols)),
    format(Sys.time(), "%d %b %Y, %H:%M")
  )
)

if (".submitted" %in% names(responses) && any(!is.na(responses$.submitted))) {
  overview <- overview |>
    add_row(
      Item = c("First response", "Latest response"),
      Value = c(
        format(min(responses$.submitted, na.rm = TRUE), "%d %b %Y, %H:%M"),
        format(max(responses$.submitted, na.rm = TRUE), "%d %b %Y, %H:%M")
      ),
      .before = 3
    )
}

kable(overview, align = c("l", "l"))
```

```{r}
#| label: responses-over-time
#| eval: !expr '".submitted" %in% names(responses) && sum(!is.na(responses$.submitted)) >= 5'
#| fig-height: 3

responses |>
  filter(!is.na(.submitted)) |>
  count(day = as_date(.submitted), name = "n") |>
  mutate(cumulative = cumsum(n)) |>
  ggplot(aes(day, cumulative)) +
  geom_line(colour = navy, linewidth = 1) +
  geom_point(colour = gold, size = 2.2) +
  scale_x_date(date_labels = "%d %b") +
  labs(x = NULL, y = "Cumulative responses", title = "Responses received over time") +
  theme_minimal(base_size = 12, base_family = "Georgia")
```

## Single-choice questions

```{r}
#| label: single-choice
#| output: asis

if (length(single_cols) == 0) {
  cat("\n_No single-choice questions detected._\n\n")
}

walk(single_cols, \(col) {
  tbl <- tally_single(responses, col)
  if (nrow(tbl) == 0) return(invisible(NULL))

  cat("\n\n### ", question_labels[[col]], "\n\n", sep = "")
  print(plot_tally(tbl, subtitle = str_glue("n = {sum(tbl$n)}")))
  cat("\n\n")
})
```

## Multi-select questions

Percentages are of respondents, so they may total more than 100%.

```{r}
#| label: multi-select
#| output: asis

if (length(multi_cols) == 0) {
  cat("\n_No multi-select questions detected._\n\n")
}

walk(multi_cols, \(col) {
  tbl <- tally_multi(responses, col)
  if (nrow(tbl) == 0) return(invisible(NULL))

  cat("\n\n### ", question_labels[[col]], "\n\n", sep = "")
  print(plot_tally(tbl, subtitle = "% of respondents selecting each option"))
  cat("\n\n")
})
```

## Open-ended responses

```{r}
#| label: free-text
#| output: asis

if (length(text_cols) == 0) {
  cat("\n_No open-ended questions detected._\n\n")
}

walk(text_cols, \(col) {
  answers <- responses[[col]] |>
    str_trim() |>
    discard(\(x) is.na(x) | x == "" | str_to_lower(x) %in% c("na", "n/a", "-", "none"))

  if (length(answers) == 0) return(invisible(NULL))

  cat("\n\n### ", question_labels[[col]], "\n\n", sep = "")
  cat("_", length(answers), " of ", n_responses, " respondents answered._\n\n", sep = "")
  walk(answers, \(a) cat("> ", str_replace_all(a, "\n", " "), "\n\n", sep = ""))
})
```

## Data audit

```{r}
#| label: audit

tibble(
  question = question_labels[candidate_cols],
  type     = recode(
    col_type[candidate_cols],
    single = "Single choice",
    multi  = "Multi-select",
    text   = "Free text",
    empty  = "No responses yet"
  ),
  answered = map_int(candidate_cols, \(v) sum(!is.na(responses[[v]]) & str_trim(responses[[v]]) != "")),
  missing  = n_responses - answered
) |>
  kable(col.names = c("Question", "Detected type", "Answered", "Blank"))
```

::: {.callout-note}
Column types above are inferred from the data, not from the form definition.
If a question is mis-classified — for example a short free-text answer read as
single choice — add its cleaned column name to `single_cols`, `multi_cols`, or
`text_cols` in the `classify-columns` chunk before rendering.
:::
