---
title: "Data Analysis Using R for Evidence-Based Decision-Making"
subtitle: "Pre-Training Survey — Results in Just Minutes"
title-block-banner: true
author: "Data for Development (D4D), School of Economics, QAU"
date: today
format:
  html:
    theme: default
    toc: true
    toc_depth: 2
    toc_float: true
    keep-md: false
    highlight-style: dracula
execute:
  warning: false
  message: false
  echo: false
  error: true
---

```{r}
#| label: setup
#| error: false
library(tidyverse)
library(googlesheets4)
library(janitor)
library(hrbrthemes)
library(scales)
library(knitr)

# ---------------------------------------------------------------------------
# AUTHENTICATION -- cached OAuth, same as the "From Survey to Insights" file
#
# gs4_auth(email=) signs in as YOU through gargle's built-in OAuth client,
# which lives in the tidyverse team's Google Cloud project. Two consequences,
# both good: the Sheets API is already enabled over there so there is nothing
# to switch on, and you are the sheet's owner so there is nothing to share.
#
# The one change from the older file is cache. cache = FALSE forces a full
# browser handshake on EVERY render -- fine when you are sitting in RStudio
# clicking through it, impossible in a background or scheduled render. Caching
# to .secrets/ means you authorise once and every render after that is silent.
#
#   FIRST RUN: a browser opens, you pick g.zahid@gmail.com, you approve.
#   EVERY RUN AFTER: nothing. It reads the cached token.
#
# .secrets/ holds a live token -- add it to .gitignore alongside the .Rproj
# junk. If the token ever goes stale (password change, revoked access), delete
# the .secrets folder and the next render re-runs the browser step once.
#
# If this report ever has to render somewhere that is NOT your laptop --
# GitHub Actions, a server, a colleague's machine -- cached OAuth cannot go
# there and you want a service account instead. That needs the Sheets API
# enabled on your own Cloud project AND the sheet shared with the service
# account's client_email. Not needed for a laptop workflow.
# ---------------------------------------------------------------------------

options(
  gargle_oauth_cache = ".secrets",
  gargle_oauth_email = "g.zahid@gmail.com"
)

gs4_auth(email = "g.zahid@gmail.com", cache = ".secrets")

sheet_url <- "https://docs.google.com/spreadsheets/d/1Ttg8BfByE9BEW7fy0YtLtWwBY7VmV7V0COdr0EgUjNc/edit?gid=1358570388#gid=1358570388"

theme_set(theme_ipsum(base_family = "", axis_title_family = ""))
navy <- "#0f2440"
gold <- "#c9a227"
blue <- "#6CABDD"
```

```{css}
body {
  font-size: 18px;
}
```

```{r}
#| label: read-data
#| error: false
survey_responses <- tryCatch(
  read_sheet(sheet_url) |> clean_names(),
  error = function(e) {
    msg <- conditionMessage(e)

    hint <- if (str_detect(msg, "PERMISSION_DENIED|403|does not have permission")) {
      paste0(
        "Signed in, but that Google account cannot see this sheet.\n",
        "Check you approved the browser prompt as g.zahid@gmail.com and not ",
        "some other signed-in account. Deleting the .secrets folder forces a ",
        "fresh sign-in."
      )
    } else if (str_detect(msg, "404|NOT_FOUND|not found")) {
      "Sheet not found -- check sheet_url still points at a live sheet."
    } else if (str_detect(msg, "oauth|token|credential|auth")) {
      paste0(
        "Auth failed. If this render is unattended it cannot open a browser: ",
        "run it once interactively to populate .secrets/, then retry."
      )
    } else {
      "Unrecognised failure; read the original error below."
    }

    stop("Could not read the sheet.\n\n", hint, "\n\nOriginal error: ", msg)
  }
)

n_responses <- nrow(survey_responses)
today_pretty <- format(Sys.Date(), "%B %d, %Y")
```

```{r}
#| label: helpers
#| error: false

# read_sheet() returns list-columns whenever a column has mixed types (a
# stray blank in a numeric question is enough). Everything below is plotted
# as a category, so flatten to character once and stop worrying about it.
# This also fixes fct_reorder(), which errors on a numeric vector.
as_chr <- function(x) {
  if (is.list(x)) {
    x <- map_chr(x, \(v) if (length(v) == 0 || is.null(v)) NA_character_ else as.character(v)[1])
  }
  str_trim(as.character(x))
}

clean_answers <- function(x) {
  x <- as_chr(x)
  discard(x, \(v) is.na(v) | v == "" | str_to_lower(v) %in% c("na", "n/a", "-", "none"))
}

# Find a column by pattern; returns NA if the question isn't in the form yet.
pick <- function(pattern) {
  hit <- str_subset(names(survey_responses), pattern)
  if (length(hit) > 0) hit[1] else NA_character_
}

# Single-choice bar chart. ordered = TRUE sorts by the answer label instead
# of by count -- right for scale questions like "1 - Beginner", "2 - ...".
bar_single <- function(col, title, fill = blue, ordered = FALSE) {
  if (is.na(col)) {
    cat("_Not asked in the form yet._\n\n")
    return(invisible(NULL))
  }
  tbl <- tibble(answer = as_chr(survey_responses[[col]])) |>
    filter(!is.na(answer), answer != "") |>
    count(answer, sort = !ordered)

  if (nrow(tbl) == 0) {
    cat("_No answers yet._\n\n")
    return(invisible(NULL))
  }

  tbl <- tbl |>
    mutate(answer = if (ordered) fct_rev(factor(answer, sort(unique(answer))))
                    else fct_reorder(answer, n))

  print(
    ggplot(tbl, aes(answer, n)) +
      geom_col(fill = fill) +
      geom_text(aes(label = n), hjust = 1.3, colour = "white", fontface = "bold") +
      coord_flip() +
      scale_y_continuous(breaks = breaks_pretty()) +
      labs(title = title, x = NULL, y = "Responses") +
      theme_minimal()
  )
}

# Multi-select ("select all that apply") -- one row per chosen option.
# Splits on commas, so an option label containing a comma would be split
# wrongly. None of the current options do; check if you add one that does.
bar_multi <- function(col, title, fill = gold) {
  if (is.na(col)) {
    cat("_Not asked in the form yet._\n\n")
    return(invisible(NULL))
  }
  tbl <- tibble(raw = as_chr(survey_responses[[col]])) |>
    filter(!is.na(raw), raw != "") |>
    mutate(answer = str_split(raw, ",\\s*")) |>
    unnest(answer) |>
    mutate(answer = str_trim(answer)) |>
    filter(answer != "") |>
    count(answer, sort = TRUE)

  if (nrow(tbl) == 0) {
    cat("_No answers yet._\n\n")
    return(invisible(NULL))
  }

  print(
    ggplot(tbl, aes(fct_reorder(answer, n), n)) +
      geom_col(fill = fill) +
      geom_text(aes(label = n), hjust = 1.3, colour = "white", fontface = "bold") +
      coord_flip() +
      scale_y_continuous(breaks = breaks_pretty()) +
      labs(title = title, x = NULL, y = "Times selected") +
      theme_minimal()
  )
}

quote_answers <- function(col) {
  if (is.na(col)) {
    cat("_Not asked in the form yet._\n\n")
    return(invisible(NULL))
  }
  answers <- clean_answers(survey_responses[[col]])
  cat("_", length(answers), " of ", n_responses, " respondents answered._\n\n", sep = "")
  walk(answers, \(a) cat("> ", str_replace_all(a, "\n", " "), "\n\n", sep = ""))
}
```

```{r}
#| label: columns
#| error: false
# Patterns matched against the real cleaned column names from this form.
# If you rename a question in Google Forms, update the pattern here --
# anything that stops matching quietly falls through to "Other Questions".
org_col     <- pick("organi[sz]ation|department|affiliation")
role_col    <- pick("job_title|designation")
exp_col     <- pick("years_of_professional_experience")
edu_col     <- pick("highest_level_of_education")
usedr_col   <- pick("have_you_used_r_before")
skill_col   <- pick("overall_r_skill|skill_level")
tools_col   <- pick("used_for_data_analysis")
worked_col  <- pick("already_worked_with")
ai_col      <- pick("ai_coding_assistant")
setup_col   <- pick("set_up_on_your_laptop")
dataset_col <- pick("bring_your_own_dataset")
comfort_col <- pick("general_computer_skills")
goals_col   <- pick("get_out_of_this_workshop|expect|hope.*learn")
help_col    <- pick("anything_specific")
notes_col   <- pick("anything_else")
```

# About Our Respondents

As of `r today_pretty`, the pre-training survey has received `r n_responses` responses.

<!--
ADDING QUESTIONS
A new question in the form becomes a new column here after clean_names().
Anything without its own section below appears automatically in "Other
Questions" at the end, so nothing is ever lost. To promote one, add a
pick() line in the `columns` chunk and a section here.
-->

## Organization

```{r}
#| label: organization
bar_single(org_col, "Organization")
```

## Job Title / Designation

```{r}
#| label: role
bar_single(role_col, "Job title / designation")
```

## Years Working With Data

```{r}
#| label: experience
bar_single(exp_col, "Years of professional experience working with data", ordered = TRUE)
```

## Highest Level of Education

```{r}
#| label: education
bar_single(edu_col, "Highest level of education completed")
```

# Where the Room Is Starting From

## Have They Used R Before?

```{r}
#| label: used-r
bar_single(usedr_col, "Have you used R before?")
```

## Self-Rated R Skill

```{r}
#| label: skill
bar_single(skill_col, "Self-rated R skill level", ordered = TRUE)
```

## Software Already Used for Analysis

```{r}
#| label: tools
bar_multi(tools_col, "Used for data analysis before")
```

## R Ecosystem Exposure

```{r}
#| label: worked-with
bar_multi(worked_col, "Already worked with")
```

## AI Coding Assistants

```{r}
#| label: ai
bar_single(ai_col, "Used AI coding assistants before?")
```

## Comfort With General Computer Skills

```{r}
#| label: comfort
bar_single(comfort_col, "Comfort installing software / managing files", ordered = TRUE)
```

# Logistics

## What's Already Installed

```{r}
#| label: setup-status
bar_multi(setup_col, "Already set up on their laptop")
```

These are the numbers that decide how Day 1 morning goes. Anyone missing R or
RStudio needs to be caught before the session, not during it.

## Bringing Their Own Data

```{r}
#| label: own-dataset
bar_single(dataset_col, "Plans to bring their own dataset")
```

# What They Want

## Priorities for the Workshop

```{r}
#| label: goals
bar_multi(goals_col, "What they most want out of the workshop (up to 2 each)")
```

## Specific Help Requested

```{r}
#| label: help
#| output: asis
quote_answers(help_col)
```

## Anything Else for the Trainers

```{r}
#| label: notes
#| output: asis
quote_answers(notes_col)
```

# Other Questions

Any question not given its own section above appears here automatically.

```{r}
#| label: other-questions
#| output: asis

named_cols <- c(org_col, role_col, exp_col, edu_col, usedr_col, skill_col,
                tools_col, worked_col, ai_col, setup_col, dataset_col,
                comfort_col, goals_col, help_col, notes_col) |>
  discard(is.na) |>
  unique()

# Identifying / admin columns are never charted -- they are personal data.
id_pattern <- "timestamp|time_stamp|submitted|email|e_mail|score|^name$|full_name|phone|cnic|contact"

remaining_cols <- setdiff(
  names(survey_responses)[!str_detect(names(survey_responses), id_pattern)],
  named_cols
)

classify_column <- function(x) {
  x <- as_chr(x)
  x <- x[!is.na(x) & x != ""]
  if (length(x) == 0) return("empty")

  n_unique   <- n_distinct(x)
  mean_len   <- mean(str_length(x))
  comma_frac <- mean(str_detect(x, ","))
  split_vals <- str_split(x, ",\\s*") |> unlist() |> str_trim() |> unique()

  if (mean_len > 60 || (n_unique >= 0.9 * length(x) && mean_len > 30)) return("text")
  if (comma_frac > 0.15 && length(split_vals) < n_unique) return("multi")
  "single"
}

if (length(remaining_cols) == 0) {
  cat("_Every question has its own section above._\n\n")
}

walk(remaining_cols, \(col) {
  type <- classify_column(survey_responses[[col]])
  if (type == "empty") return(invisible(NULL))

  heading <- str_to_sentence(str_replace_all(col, "_", " "))
  cat("\n\n## ", heading, "\n\n", sep = "")

  if (type == "text")       quote_answers(col)
  else if (type == "multi") bar_multi(col, heading)
  else                      bar_single(col, heading)
})
```
